% !TEX TS-program = pdflatex
% Compile (pdfLaTeX):
%   pdflatex "Audit-Closed AI Scientist Protocol.tex"
%   pdflatex "Audit-Closed AI Scientist Protocol.tex"

\documentclass[11pt,a4paper]{article}

% ---------- Machine-readability / PDF text extraction ----------
\usepackage{cmap}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\IfFileExists{glyphtounicode.tex}{\input{glyphtounicode.tex}\pdfgentounicode=1}{}
\usepackage{microtype}
\usepackage{setspace}
\setstretch{1.15}
\usepackage{float}

% ---------- Math / theorem ----------
\usepackage{amsmath,amssymb,amsthm,mathtools,bm}
\usepackage{bbm}

% ---------- Layout / utils ----------
\usepackage[margin=0.8in]{geometry}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{url}
\usepackage{array}
\usepackage{booktabs}
\usepackage{verbatim}

% ---------- Algorithms ----------
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

% ---------- Citations ----------
\usepackage[numbers,sort&compress]{natbib}

% ---------- Hyperref metadata (hyperref only; no hyperxmp) ----------
\usepackage[
  pdftitle={Audit-Closed AI Scientist Protocol: Trustworthy Autonomous Scientific Discovery for Self-Driving Laboratories},
  pdfauthor={K. Takahashi},
  pdfsubject={agentic AI scientist governance; autonomous research agents; auditable theory and measurement expansion},
  pdfkeywords={AI Scientist, agentic AI scientist, autonomous scientific discovery, open-ended scientific discovery, autonomous research agents, self-driving laboratories, categorical semantics, e-values, optional stopping, active learning, drift localization, reproducibility},
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
]{hyperref}

% ---------- Hyphenation / line breaks ----------
\hyphenpenalty=800
\exhyphenpenalty=800
\sloppy

% ---------- Theorem environments ----------
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}

\newtheorem{definition}{Definition}[section]
\newtheorem{assumption}{Assumption}[section]
\newtheorem{remark}{Remark}[section]

% ---------- Convenience macros ----------
\newcommand{\Obs}{\mathbf{Obs}}
\newcommand{\Stoch}{\mathbf{Stoch}}
\newcommand{\Th}{\mathbf{Th}}
\newcommand{\Log}{\mathsf{Log}}
\newcommand{\Cert}{\mathsf{Cert}}
\newcommand{\Gen}{\mathsf{Genesis}}
\newcommand{\Acc}{\mathsf{Accept}}
\newcommand{\Rej}{\mathsf{Reject}}
\newcommand{\Quar}{\mathsf{Quarantine}}
\newcommand{\Prov}{\mathsf{Provisional}}
\newcommand{\Adopt}{\mathsf{Adopt}}
\newcommand{\Score}{\mathsf{Score}}
\newcommand{\Cost}{\mathsf{Cost}}
\newcommand{\Cand}{\mathsf{Cand}}
\newcommand{\TV}{\mathsf{TV}}
\newcommand{\KL}{\mathsf{KL}}
\newcommand{\MI}{\mathsf{I}}
\newcommand{\EvScore}{\mathsf{EvScore}}

\title{\textbf{Audit-Closed AI Scientist Protocol:}\\
\Large{Trustworthy Autonomous Scientific Discovery for Self-Driving Laboratories}}
\author{K. Takahashi\\
\footnotesize \href{https://orcid.org/0009-0004-4273-3365}{ORCID: 0009-0004-4273-3365}}
\date{February 22, 2026}

\begin{document}
\maketitle

\begin{abstract}
AI scientist systems and self-driving laboratories require governance protocols that are statistically valid under continuous monitoring, auditable without privileged operators, and robust to physical-world failure modes.
We present an audit-closed protocol in which every decision is a deterministic function of a public transparency log, with all semantic claims grounded in explicitly declared observable contexts.

The formal core integrates typed stochastic observation interfaces, tolerance-aware predictive semantics, and non-vacuous coherence auditing on finite acyclic proof graphs.
To resolve the deterministic-pipeline tension, we combine explicit deterministic-depth controls with a Lipschitz observation-translation theorem that converts latent Wasserstein closeness into observable total-variation guarantees.

For sequential inference we use always-valid e-processes with gate-wise alpha accounting, robust score sanitization plus integrity monitoring, and anti-griefing novelty dynamics with signal-qualified eligible-cost updates.
For adaptive experimentation we provide two regimes: a conservative fixed-ratio IPS test and a variance-adaptive predictable betting process that preserves optional-stopping validity while avoiding high-ratio evidence paralysis.
For physical deployment we introduce hierarchical primary/secondary sentinels, spoofing-aware coherence gates, and auditable calibration transitions that separate sensor aging from sentinel degradation.

For long-horizon autonomy we use drift-mode state transitions, exogenous-only root-memory decay, and subgraph-local fault localization with familywise error control.
We prove system-level guarantees covering adaptive optional stopping, variance-adaptive evidence growth, Byzantine-resilient aggregation, legal-spam stability, drift-localization safety, and mixture re-anchoring coherence.
A fully reproducible synthetic benchmark confirms improved availability under outliers, high-ratio contextual exploration, root-debt erosion attacks, and sentinel aging while maintaining conservative false-acceptance behavior; the released manifest includes cryptographic digests and an automatic manuscript-value consistency check.
\end{abstract}

\noindent\textbf{Keywords:} AI Scientist; agentic AI scientist; autonomous scientific discovery; open-ended scientific discovery; autonomous research agents; self-driving laboratories; e-values; optional stopping; active learning; drift localization; reproducibility.

\section{Introduction}

Equation discovery recovers compact governing equations from fixed observables \citep{SchmidtLipson2009,UdrescuTegmark2020,Brunton2016SINDy}.
Scientific machine learning integrates mechanistic structure with flexible approximation \citep{Raissi2019PINN,Karniadakis2021PIML,Cuomo2022PINNReview}.
Robot Scientists and self-driving laboratories automate closed-loop experimentation \citep{King2009Automation,Sparkes2010RobotScientists,Tom2024SDLchemrev,TobiasWahab2025SDLReview}.
Recent agentic AI scientist systems target end-to-end automation of idea generation, experiment execution, and report writing \citep{Lu2024AIScientist,Yamada2025AIScientistV2}.
Bayesian optimal experimental design (BOED) formalizes experiment selection \citep{Lindley1956,ChalonerVerdinelli1995,Foster2019vOED}, with amortized approaches such as iDAD \citep{Ivanova2021iDAD}.

A persistent gap is \emph{epistemic governance under physical constraints}: how to justify updates without privileged adjudicators while confronting noise, outliers, high-dimensional observables, adversarial behavior, and finite lab budgets.
This paper adopts fallibilism \citep{Popper1959,Quine1951} and treats progress as testabilization under audit constraints: new observables are valuable only when they produce replayable and sequentially valid evidence.

\paragraph{Non-claims.}
No claim is made that the protocol discovers all laws or removes the need for initial human commitments.
All guarantees are conditional on explicit, logged Genesis commitments and stated trust assumptions.

\section{Threat Model and Failure Modes}

We treat the following failure modes as first-class constraints:
\begin{itemize}[leftmargin=*,itemsep=2pt]
\item \textbf{F1: Optional stopping and peeking.} Continuous monitoring invalidates naive tests \citep{Johari2017Peeking,Johari2022AlwaysValid}.
\item \textbf{F2: Many-candidate inflation.} Trying many candidates without correction yields spurious ``wins''.
\item \textbf{F3: Candidate shopping.} Unlogged candidate search hides failures behind a winner.
\item \textbf{F4: Data-role leakage.} Reusing evaluation data for fitting or selection invalidates evidence.
\item \textbf{F5: Non-replayable execution.} Missing provenance or nondeterminism prevents verification \citep{Peng2011,Sandve2013}.
\item \textbf{F6: Infrastructure privilege.} In multi-agent settings, hidden repositories or privileged submitters create structural advantage.
\item \textbf{F7: Oracle boundary attacks.} Digitally logged data may be fabricated and not physically measured.
\item \textbf{F8: Sybil attacks.} Adversaries may create many pseudo-agents to skew federated evidence.
\item \textbf{F9: Outlier denial-of-service.} Rare sensor glitches can repeatedly invalidate strict score transcripts.
\item \textbf{F10: Safety-induced paralysis.} Overly strict conjunction gates can block all useful updates.
\item \textbf{F11: Adaptive-sampling mismatch.} Contextual experiment selection can invalidate fixed-distribution score interpretation unless logged propensity correction is built into inference.
\item \textbf{F12: Sentinel ambiguity.} Apparent calibration drift can originate from sensor aging or sentinel aging; single-sentinel schemes cannot separate them.
\item \textbf{F13: Root-memory erosion attack.} Repeated induced drift triggers can force debt-decay schedules to forget historical anchors unless decay is tied to exogenous physical evidence.
\end{itemize}

\section{Audit-Closed Governance, Provenance, and State Transitions}

\subsection{Audit-closed update rule}
\begin{definition}[Audit-closed update rule]
Let $\Log_{0:t}$ be the public log prefix up to time $t$.
An update rule is \emph{audit-closed} if there exists a deterministic function $f$ such that
\[
\Acc_t=f(\Log_{0:t}).
\]
\end{definition}

\begin{definition}[Fail-closed and fail-soft containment]
Critical verification failures (missing provenance, invalid signatures, undefined transcript steps, or budget overflows) force $\Rej$.
Non-critical witness failures are mapped to neutral evidence contributions and audited separately; they cannot increase adoption probability.
\end{definition}

\subsection{Tamper-evident public logs}
\begin{definition}[Transparency log]
A transparency log is an append-only log committed by signed Merkle-tree roots.
Verifiers obtain inclusion and consistency proofs \citep{RFC6962}.
\end{definition}

To mitigate equivocation, Genesis requires witness cosigning and gossip \citep{Chuat2015Gossip,Oxford2020CTGossip}.

\subsection{Physical provenance boundary}
\begin{definition}[Attested measurement record]
Each measurement record includes:
device ID, firmware digest, calibration-certificate digest, protocol digest, challenge nonce, timestamp, payload digest, and device signature.
Challenge nonces are derived from a public randomness beacon \citep{NISTIR8213draft,NISTBeaconProject}.
\end{definition}

\begin{assumption}[Minimal physical trust base]
At least one attested sensing path per accepted certificate is uncompromised and verifiable.
All claims about physical evidence are conditional on this assumption.
\end{assumption}

\begin{definition}[Cross-sensor physical coherence witnesses]
Genesis specifies a finite set of physical-consistency witnesses
\[
\mathcal{G}_{\mathrm{phys}}=\{g_1,\dots,g_M\},
\]
where each $g_m$ is a deterministic function of synchronized multi-sensor records and physically interpretable metadata (e.g., conservation constraints, instrument cross-calibration identities, monotonicity laws).
\end{definition}

\begin{definition}[Physical spoofing null]
For each witness $g_m$, define residual $R_{m,i}$ on record $i$ and test
\[
H^{\mathrm{phys}}_{0,m}:\ \mathbb{E}[R_{m,i}\mid \mathcal{F}_{i-1}]\le \tau^{\mathrm{phys}}_m
\]
using an always-valid e-process on attested synchronized streams.
\end{definition}

\begin{remark}
Attestation proves digital provenance, not physical honesty of the environment.
Physical spoofing resistance is therefore enforced by explicit multi-sensor coherence tests.
\end{remark}

\begin{definition}[Hierarchical sentinel architecture]
Genesis defines two sentinel layers:
\begin{itemize}[leftmargin=*]
\item \textbf{Primary sentinels} $\mathcal{C}_{\mathrm{cal}}^{(P)}$: slow-changing references with stricter custody and replacement governance.
\item \textbf{Secondary sentinels} $\mathcal{C}_{\mathrm{cal}}^{(S)}$: operational references that may age and are periodically recalibrated against primaries.
\end{itemize}
Calibration certificates must indicate which layer supported each decision.
\end{definition}

\begin{definition}[Calibration-update candidate with hierarchy checks]
A calibration candidate proposes
\[
(\theta^{\mathrm{cal}}_{\mathrm{old}}\to \theta^{\mathrm{cal}}_{\mathrm{new}},\ \tau^{\mathrm{phys}}_m\to \tau^{\mathrm{phys}\,'}_m),
\]
and is admissible only if:
\begin{enumerate}[leftmargin=*]
\item primary sentinel health checks pass on fresh attested holdout runs;
\item secondary-vs-primary consistency checks identify whether drift is sensor-side or sentinel-side;
\item spoofing indicators remain below Genesis limits during recalibration runs; and
\item post-update physical witnesses pass on fresh holdout data disjoint from candidate design runs.
\end{enumerate}
\end{definition}

\begin{proposition}[Sentinel-collapse prevention rule]
If primary health checks fail, sensor recalibration is forbidden in that epoch and the system enters a sentinel-maintenance state.
This prevents the feedback loop ``degraded sentinel $\to$ wrong sensor recalibration $\to$ amplified inconsistency''.
\end{proposition}

\subsection{Update states and quarantine entry}
\begin{definition}[Update states]
A candidate update has state $\Rej$, $\Quar$, $\Prov$, or $\Adopt$.
Quarantine permits bounded pilot execution.
Provisional state permits bounded-scope deployment while stronger evidence is accumulated.
\end{definition}

\begin{definition}[Quarantine entry conditions]
A candidate may enter $\Quar$ only if it passes (i) static safety prechecks (hazard class, resource caps), (ii) candidate-set commitment and generator admissibility checks, and (iii) a quarantine budget check (max pilot trials, max energy/material limits), all fixed in Genesis and verifiable from the log.
\end{definition}

\begin{remark}
Calibration-update candidates are treated as restricted-scope candidates: they may update hardware calibration parameters and physical-gate baselines, but cannot change theory content in the same certificate.
\end{remark}

\subsection{Pilot-data non-leakage contract}
\begin{definition}[Permitted summary algebra]
Genesis fixes a finite set $\mathcal{S}_{\mathrm{pilot}}=\{s_1,\dots,s_M\}$ of deterministic summary functions on pilot data and a closure rule under \emph{non-adaptive} post-processing.
The candidate generator may depend on pilot data only through $(s_1(D_{\mathrm{P}}),\dots,s_M(D_{\mathrm{P}}))$, and only for budgeting (e.g., continue/stop pilot), not for proposing new candidate content.
\end{definition}

\begin{definition}[Strict non-reuse of pilot data]
Pilot data $D_{\mathrm{P},t}$ is never reused as conservativity/progress evidence data.
\end{definition}

\section{Typed Observable Semantics with Non-Vacuous Coherence}

This section addresses four constraints simultaneously:
(i) typed context-dependent outcome spaces,
(ii) non-vacuous coherence control under long pipelines,
(iii) comparability without universal pullback assumptions, and
(iv) operational checkability from logged artifacts.

\subsection{The category of stochastic maps}
\begin{definition}[Stochastic map category $\Stoch$]
$\Stoch$ has objects measurable spaces $(X,\Sigma_X)$ and morphisms Markov kernels $K:X\rightsquigarrow Y$.
Composition is kernel composition; identities are Dirac kernels.
\end{definition}

\subsection{Total variation and contraction}
\begin{definition}[Total variation]
For measures $\mu,\nu$ on $(X,\Sigma)$,
$\TV(\mu,\nu)=\sup_{A\in\Sigma}|\mu(A)-\nu(A)|$.
\end{definition}

\begin{definition}[Dobrushin contraction coefficient]
For a kernel $K:X\rightsquigarrow Y$, define
\[
\eta(K)=\sup_{x,x'\in X}\TV\bigl(K(x,\cdot),K(x',\cdot)\bigr)\in[0,1].
\]
\end{definition}

\begin{remark}[Deterministic-map caveat]
For deterministic measurable maps represented as Dirac kernels, typically $\eta(K)=1$.
Hence deterministic processing chains do not contract TV by themselves and must be controlled by explicit depth/budget constraints or by a certified metric-translation argument at the observation interface.
\end{remark}

\begin{lemma}[Kernel contraction with coefficient]
For a Markov kernel $K:X\rightsquigarrow Y$ and measures $\mu,\nu$ on $X$,
\[
\TV(K_\#\mu,K_\#\nu)\le \eta(K)\TV(\mu,\nu)\le \TV(\mu,\nu).
\]
\end{lemma}

\subsection{Predictions as a \emph{budgeted} toleranced category}
\begin{definition}[$\mathsf{Pred}^{\TV}$]
Objects are pairs $(X,\mu)$ with $X\in\mathrm{Ob}(\Stoch)$ and $\mu$ a probability measure on $X$.
A morphism $(X,\mu)\to(Y,\nu)$ is a pair $(K,\epsilon)$ where $K:X\rightsquigarrow Y$ is a Markov kernel and $\epsilon\ge 0$ satisfies
\[
\TV(K_\#\mu,\nu)\le \epsilon.
\]
Composition is $(L,\epsilon')\circ(K,\epsilon)=(L\circ K,\epsilon+\epsilon')$.
Identity is $(\mathrm{id},0)$.
\end{definition}


\begin{remark}[Preorder interpretation of tolerances]
We interpret $\epsilon$ as an \emph{upper bound} on inconsistency.
For a fixed kernel $K$, a smaller tolerance is strictly stronger.
Accordingly, we equip each hom-set with a preorder:
\[
(K,\epsilon)\preceq (K,\epsilon')\quad \Longleftrightarrow\quad \epsilon\le \epsilon',
\]
making $\mathsf{Pred}^{\TV}$ a \emph{locally preordered} category.
This supports \emph{lax} commutativity conditions where required.
\end{remark}

\begin{remark}
Unlike strict element categories requiring $K_\#\mu=\nu$, $\mathsf{Pred}^{\TV}$ permits controlled incoherence.
To prevent vacuity, Genesis must bound tolerance accumulation (Definition~\ref{def:cohbudget}).
\end{remark}

\begin{definition}[Audit expression graph and contracted coherence budgets]
\label{def:cohbudget}
For each certificate, Genesis specifies a finite \emph{acyclic audit expression graph} $G_t$.
Its edges are exactly the morphisms used in audit predicates and derived quantities.
Each edge $u$ has:
\begin{itemize}[leftmargin=*]
\item a tolerance budget $b(u)\in[0,\infty)$;
\item a logged contraction upper bound $\bar\eta(u)\in[0,1]$.
\end{itemize}

For a directed path $p=(u_1,\dots,u_\ell)$ in $G_t$, define the contracted path budget
\[
B(p)=\sum_{j=1}^{\ell}\left(\prod_{k=j+1}^{\ell}\bar\eta(u_k)\right)b(u_j).
\]
Define deterministic-depth and deterministic-budget terms
\[
d_{\det}(p)=\sum_{j=1}^{\ell}\mathbbm{1}\{\bar\eta(u_j)=1\},
\qquad
B_{\det}(p)=\sum_{j:\bar\eta(u_j)=1} b(u_j).
\]
A theory is coherence-admissible in epoch $t$ if
\[
\max_{p\in\mathrm{Path}(G_t)}B(p)\le \epsilon_{\mathrm{coh}}<1,
\]
and, for Genesis constants $L_{\det}$ and $\epsilon_{\det}<1$,
\[
\max_{p\in\mathrm{Path}(G_t)}d_{\det}(p)\le L_{\det},
\qquad
\max_{p\in\mathrm{Path}(G_t)}B_{\det}(p)\le \epsilon_{\det}.
\]
\end{definition}

\begin{proposition}[Non-vacuity of certified coherence bounds]
If Definition~\ref{def:cohbudget} holds, then every audit-derived inconsistency bound in the certificate is strictly below $1$ and therefore non-vacuous.
\end{proposition}
\begin{proof}[Proof sketch]
Apply the contraction lemma along each path in $G_t$.
For $\bar\eta<1$ segments, contraction-weighted accumulation controls growth.
For $\bar\eta=1$ segments, deterministic-depth and deterministic-budget constraints prevent vacuous additive blow-up.
Acyclicity prevents repeated-cycle blow-up.
\end{proof}

\begin{corollary}[Deterministic-chain regime]
If all edges on a path satisfy $\bar\eta=1$, then $B(p)=B_{\det}(p)$.
Hence non-vacuity is guaranteed either through explicit deterministic-depth/deterministic-budget constraints or through a certified metric-translation route at the observation interface (Theorem~\ref{thm:tv-w1}).
\end{corollary}

\begin{definition}[Audited bottleneck edge]
An edge $u_b$ in $G_t$ is an \emph{audited bottleneck} if:
\begin{enumerate}[leftmargin=*]
\item $\bar\eta(u_b)\le \eta_b<1$ is certified;
\item $u_b$ maps to a lower-complexity context (e.g., sufficient statistic or conservation-law summary) with explicit replay code; and
\item either (a) an induced stochastic contraction contract establishing $\eta_b<1$, or (b) a downstream Lipschitz-smoothed observation translation certificate (Theorem~\ref{thm:tv-w1}) is logged.
\end{enumerate}
\end{definition}

\begin{theorem}[Topological budget compression via information bottlenecks]
\label{thm:bottleneck}
Let a path decompose as $p=p_{\mathrm{pre}}\cdot u_b\cdot p_{\mathrm{post}}$, where $u_b$ is an audited bottleneck with coefficient $\eta_b<1$.
Define
\[
B_{\mathrm{pre}}=\sum_{u\in p_{\mathrm{pre}}}b(u),
\qquad
B_{\mathrm{post}}=\sum_{u\in p_{\mathrm{post}}}b(u).
\]
Then the certified path budget satisfies
\[
B(p)\le B_{\mathrm{pre}}+b(u_b)+\eta_b\,B_{\mathrm{post}}.
\]
For an otherwise identical path with no contracting bottleneck ($\eta_b=1$), the bound is
\[
B_{\mathrm{det}}(p)=B_{\mathrm{pre}}+b(u_b)+B_{\mathrm{post}}.
\]
Hence the bottleneck contracts all downstream tolerance contribution by factor $\eta_b$, enabling deeper non-vacuous audit paths.
\end{theorem}
\begin{proof}
Apply Definition~\ref{def:cohbudget}.
All terms upstream of $u_b$ are unaffected.
Every downstream edge budget is multiplied by at least one factor $\bar\eta(u_b)\le\eta_b$, giving the stated upper bound.
Setting $\eta_b=1$ recovers the deterministic-chain expression.
\end{proof}

\begin{definition}[Lipschitz-smoothed observation interface]
Let $(Z,d_Z)$ be a latent metric space and let $K_{\mathrm{obs}}:Z\rightsquigarrow Y$ be the final observation kernel.
The interface is \emph{Lipschitz-smoothed} with constant $L_{\mathrm{obs}}$ if for all probability measures $\mu,\nu$ on $Z$,
\[
\TV\!\left((K_{\mathrm{obs}})_\#\mu,(K_{\mathrm{obs}})_\#\nu\right)\le L_{\mathrm{obs}}\,W_1(\mu,\nu),
\]
where $W_1$ is the 1-Wasserstein distance on $(Z,d_Z)$ \citep[see][for foundations]{Villani2009OT}.
\end{definition}

\begin{theorem}[TV--Wasserstein translation for deterministic scientific maps]
\label{thm:tv-w1}
Let $f:X\to Z$ be deterministic and $L_f$-Lipschitz from $(X,d_X)$ to $(Z,d_Z)$.
Suppose the final observation interface on $Z$ is Lipschitz-smoothed with constant $L_{\mathrm{obs}}$.
For two candidate predictive input laws $\mu_1,\mu_2$ on $X$, define observable laws
\[
P_k=(K_{\mathrm{obs}})_\#(f_\#\mu_k),\qquad k\in\{1,2\}.
\]
Then
\[
\TV(P_1,P_2)\le L_{\mathrm{obs}}\,W_1(f_\#\mu_1,f_\#\mu_2)\le L_{\mathrm{obs}}L_f\,W_1(\mu_1,\mu_2).
\]
Consequently, deterministic abstraction chains can remain coherence-admissible without injected artificial noise when a certified Wasserstein-to-TV translation bound is logged at the observation interface.
\end{theorem}
\begin{proof}
Apply the Lipschitz-smoothed interface definition to $f_\#\mu_1,f_\#\mu_2$, then use standard Lipschitz pushforward contraction:
$W_1(f_\#\mu_1,f_\#\mu_2)\le L_f W_1(\mu_1,\mu_2)$ \citep[Ch.~6]{Villani2009OT}.
\end{proof}


\subsection{Observation interfaces and explicit refinement witnesses}
\begin{definition}[Observation interface]
An observation interface is a pair $(\Obs,\mathcal{Y})$ where $\Obs$ is a category of measurement contexts and $\mathcal{Y}:\Obs\to\Stoch$ assigns each context its outcome space and each context morphism a \emph{declared post-processing kernel}.
Crucially, $\Obs$ is a \emph{reduction category}: primitive morphisms are exactly logged post-processing maps.
Refinement claims are represented by explicit joint-context witnesses, not by primitive coarse$\to$fine arrows.
\end{definition}

\begin{definition}[Joint-context witness]
A refinement/comparability witness between contexts $o,o'$ is a logged tuple $(m,a,b)$ with span
$o \xleftarrow{a} m \xrightarrow{b} o'$ in $\Obs$.
The tuple includes executable acquisition and post-processing specifications.
\end{definition}

\begin{remark}
This makes ``fine vs coarse'' claims checkable without assuming pullbacks exist in a broad ambient category.
Only explicitly logged witnesses may be used in audit predicates.
\end{remark}

\begin{definition}[Witness library]
For each epoch, Genesis fixes a finite witness library $\mathcal{W}_t$ of allowed tuples $(m,a,b)$ and associated check procedures.
Chaining of comparability claims is permitted only when an explicit chained witness is itself in $\mathcal{W}_t$.
\end{definition}

\begin{remark}
This avoids assuming universal composability of spans.
If a needed joint witness is not present in $\mathcal{W}_t$, the corresponding comparative predicate is undefined and therefore fail-closed.
\end{remark}
\begin{definition}[Theory comparison along a span]
Given a witness $(m,a,b)$ and a theory $T$, predicted distributions on $o$ and $o'$ are obtained by pushing forward $\mu_T(m)$ along $\mathcal{Y}(a)$ and $\mathcal{Y}(b)$ within logged tolerances.
Comparability is therefore an explicit auditable claim tied to concrete measurement design.
\end{definition}

\subsection{Theories and theory morphisms}
\begin{definition}[Theory]
Given $(\Obs,\mathcal{Y})$, a theory is a functor $P_T:\Obs\to \mathsf{Pred}^{\TV}$ such that for every $u:o\to o'$, the kernel component of $P_T(u)$ equals $\mathcal{Y}(u)$.
Thus $P_T(o)=(\mathcal{Y}(o),\mu_T(o))$ and $P_T(u)=(\mathcal{Y}(u),\epsilon_T(u))$ with
$\TV(\mathcal{Y}(u)_\#\mu_T(o),\mu_T(o'))\le \epsilon_T(u)$.
\end{definition}

\begin{definition}[Lax theory category $\Th_{\mathrm{lax}}(\Obs)$]
Objects are theories.
Morphisms are \emph{lax} natural transformations in the local preorder of $\mathsf{Pred}^{\TV}$.
\end{definition}

\begin{proposition}[Sufficient condition for a theory morphism]
Let $T,T'$ be theories. Suppose there exists a family of tolerances $\delta_o\ge 0$ such that:
\begin{enumerate}[leftmargin=*]
\item (\emph{componentwise feasibility}) for every context $o$,
\[
\TV\big(\mu_T(o),\mu_{T'}(o)\big)\le \delta_o,
\]
so that $\eta_o=(\mathrm{id}_{\mathcal{Y}(o)},\delta_o)$ is a morphism in $\mathsf{Pred}^{\TV}$; and
\item (\emph{lax naturality}) for each context morphism $u:o\to o'$,
\[
\delta_o+\epsilon_{T'}(u)\le \epsilon_T(u)+\delta_{o'}.
\]
\end{enumerate}
Then the components $\eta_o=(\mathrm{id}_{\mathcal{Y}(o)},\delta_o)$ define a lax natural transformation $\eta:P_T\Rightarrow P_{T'}$ in $\Th_{\mathrm{lax}}(\Obs)$.
\end{proposition}
\begin{proof}[Proof sketch]
Condition (1) ensures each component $\eta_o$ is a well-typed morphism in $\mathsf{Pred}^{\TV}$.
Condition (2) is exactly the preorder (lax) commutativity requirement
$P_{T'}(u)\circ \eta_o \preceq \eta_{o'}\circ P_T(u)$
under the additive composition rule for tolerances, with kernel components fixed to $\mathcal{Y}(u)$.
\end{proof}

\begin{remark}
Genesis can enforce this dominance condition when theory morphisms are required (e.g., abstraction layers).
\end{remark}

\subsection{Restriction/expansion and typed audit predicates}
\begin{definition}[Observation expansion]
An expansion is a full and faithful functor $i:\Obs_0\hookrightarrow\Obs_1$ such that $\mathcal{Y}_1\circ i=\mathcal{Y}_0$.
Restriction of theories is $i^*(P_T)=P_T\circ i$.
\end{definition}

\begin{definition}[Typed audit predicates]
An audit predicate $\varphi$ is \emph{typed} by the signature it may reference:
\begin{itemize}[leftmargin=*]
\item \emph{Local} predicates $\varphi:\Sigma\to\{\mathrm{true,false}\}$ reference only contexts in a single interface $(\Obs,\mathcal{Y})$.
\item \emph{Comparative} predicates reference two interfaces linked by a specific expansion $i$ (e.g., novelty comparing $\Obs_0$ and $\Obs_1$).
\end{itemize}
\end{definition}

\begin{definition}[Reduct for local predicates]
For an expansion $i:\Obs_0\hookrightarrow\Obs_1$ and a \emph{local} predicate $\varphi$ typed over $\Obs_0$, define $i^{-1}(\varphi)=\varphi$.
For local predicates typed over $\Obs_1$, define $i^{-1}(\varphi)$ by restricting the log inputs of $\varphi$ to records mentioning only $i(\Obs_0)$-contexts.
Reduct is \emph{not defined} for comparative predicates.
\end{definition}

\begin{definition}[Audit satisfaction condition (local predicates only)]
For every expansion $i$ and every \emph{local} predicate $\varphi$ typed over $\Obs_0$,
\[
T_1\models_{\Log}\varphi \quad \Longleftrightarrow\quad i^*(T_1)\models_{\Log} i^{-1}(\varphi).
\]
\end{definition}

\begin{remark}
This avoids an ill-typed application of institution satisfaction to mixed predicates while retaining the intended ``restriction commutes with checking'' for local properties \citep{GoguenBurstall1992}.
\end{remark}

\section{Acceptance Criteria}

\subsection{Data-role partitions}
\begin{definition}[Pooled evaluation stream and role scheduler]
At epoch $t$, experiments produce an attested pooled stream $D_{\mathrm{eval},t}$.
Genesis fixes a deterministic role scheduler
\[
\rho_t:D_{\mathrm{eval},t}\to\{\mathrm{fit},A,H,P,G\}\times\{1,\dots,K\},
\]
committed before outcomes are observed.
Role partitions may be disjoint or cross-fitted; either way the assignment is replayable from the log.
\end{definition}

\begin{remark}
This supports data economy under lab constraints.
Validity relies on role assignment being predictable from committed randomness and not outcome-adaptive.
\end{remark}

\begin{definition}[Logged adaptive context policy]
For each evaluation trial $i$, the controller may choose context $X_i$ adaptively from a logged policy
\[
\pi_i(\cdot\mid\mathcal{F}_{i-1}),
\]
provided the policy (or its committed seed/state) is logged \emph{before} observing the current outcome.
Genesis enforces positivity and bounded-ratio contracts:
\[
\pi_i(x\mid\mathcal{F}_{i-1})\ge \pi_{\min}>0,\qquad
\frac{\mathcal{Q}_P(x)}{\pi_i(x\mid\mathcal{F}_{i-1})}\le c_w<\infty
\]
for all contexts in the support of the progress reference $\mathcal{Q}_P$.
The realized ratio
\[
w_i=\frac{\mathcal{Q}_P(X_i)}{\pi_i(X_i\mid\mathcal{F}_{i-1})}
\]
is logged before observing trial-$i$ outcomes and may be used in predictable betting rules.
\end{definition}

\subsection{Committed and admissible candidate generation}
\begin{definition}[Committed candidate set]
At epoch $t$, the protocol appends a commitment $\Cand_t$ defining a finite candidate set, together with generator digests, seeds/commitments, and allowed inputs.
\end{definition}

\begin{assumption}[Generator admissibility]
The generator may access only $\Log_{0:t-1}$ and fit-role data.
It must not access current-epoch evaluation-role outcomes except through Genesis-permitted pilot summaries that do not affect candidate content.
\end{assumption}

\subsection{Anchored conservativity as ``not-worse-than''}
The conservativity gate is distinct from novelty: it prevents regress on established anchors.

\begin{definition}[Anchor generator]
Genesis fixes a distribution $\mathcal{Q}_A$ over a fixed set of anchor contexts and their sampling procedures.
At epoch $t$, anchor-role records are those with role tag $(A,\cdot)$ under $\rho_t$.
\end{definition}

\begin{definition}[Conservativity null]
For candidate $j$ at epoch $t$, let
\[
\Delta_{A,t}^{(j)}=\mathcal{S}(T_j';Y)-\mathcal{S}(T;Y),
\]
where higher is better.
For tolerance $\varepsilon_{t,j}\ge 0$, the conservativity requirement is
\[
\mathbb{E}_{\mathcal{Q}_A}[\Delta_{A,t}^{(j)}]\ge -\varepsilon_{t,j}.
\]
The bad-event null is
\[
H^{\mathrm{cons}}_{0,t,j}:\ \mathbb{E}_{\mathcal{Q}_A}[\Delta_{A,t}^{(j)}]\le -\varepsilon_{t,j}.
\]
\end{definition}

\begin{theorem}[Global conservativity guarantee over epochs and candidates]
\label{thm:conserv}
Suppose candidate $j$ at epoch $t$ can pass conservativity only if an e-process rejects
$H^{\mathrm{cons}}_{0,t,j}$ at level $\alpha^{\mathrm{cons}}_{t,j}$.
If
\[
\sum_t\sum_{j\in\Cand_t}\alpha^{\mathrm{cons}}_{t,j}\le \alpha_{\mathrm{cons}},
\]
then with probability at least $1-\alpha_{\mathrm{cons}}$, no accepted candidate violates conservativity at acceptance time.
\end{theorem}
\begin{proof}[Proof sketch]
Apply Ville's inequality candidate-wise and union bound over all $(t,j)$.
\end{proof}

\subsection{Novelty gate without circularity}
\begin{definition}[Quarantined pilot execution]
Given a proposed expansion $i:\Obs_0\hookrightarrow\Obs_1$, the protocol may execute a bounded pilot budget in contexts in $\Obs_1\setminus i(\Obs_0)$ while the candidate is in $\Quar$.
Pilot-role records are logged and used only for novelty estimation, safety checks, and garbling certificates.
\end{definition}
\begin{remark}
When feasible, pilot design includes a joint-context witness $(m,a,b)$ yielding both old and new outcome views.
This ties novelty estimation to explicit comparability evidence.
\end{remark}

\begin{definition}[Resource-normalized novelty increment]
Let $N_i$ be bounded novelty gain and $C_i$ bounded incremental cost.
Define utility increment
\[
U_i=N_i-\nu_t C_i,
\]
where $\nu_t\ge0$ is a deterministic dual variable updated by
\[
\nu_{t+1}=\Pi_{[0,\nu_{\max}]}\!\left(\nu_t+\eta_\nu\big(\mathrm{clip}(\bar C_t^{\mathrm{elig}},c_{\min},c_{\max})-B_t\big)\right).
\]
Here $\bar C_t^{\mathrm{elig}}$ is the mean cost over \emph{eligible} candidates only, $B_t$ is the budget target, and $\Pi_{[0,\nu_{\max}]}$ is projection to a fixed interval.
\end{definition}

\begin{definition}[Eligible-cost accounting and anti-griefing slashing]
Genesis defines a pre-utility quality predicate $Q_{t,j}$ requiring:
\begin{enumerate}[leftmargin=*]
\item safety/provenance/conservativity admissibility; and
\item signal qualification $S_{t,j}\ge \tau_{\mathrm{signal}}$, where $S_{t,j}$ is a Genesis-fixed non-circular signal statistic (e.g., provisional log-e slope or pilot predictive gain) computed from data not reused by final acceptance gates.
\end{enumerate}
Only candidates with $Q_{t,j}=1$ contribute to $\bar C_t^{\mathrm{elig}}$.
Candidates with $Q_{t,j}=0$ are excluded from dual updates and may incur stake slashing; their cost cannot raise $\nu_t$.
\end{definition}

\begin{remark}
This blocks ``legal spam'' in which formally safe but information-free expensive proposals are injected to inflate $\nu_t$ and suppress honest novelty.
\end{remark}

\begin{definition}[Novelty utility null]
For candidate $j$ in epoch $t$, test
\[
H^{\mathrm{nov}}_{0,t,j}:\ \mathbb{E}[U_i\mid\mathcal{F}_{i-1}]\le0
\]
with an always-valid e-process at level $\alpha^{\mathrm{nov}}_{t,j}$.
\end{definition}

\begin{theorem}[Self-stabilizing novelty utility under legal-spam pressure]
\label{thm:novelty-stable}
Assume:
\begin{enumerate}[leftmargin=*]
\item dual updates use projected dynamics with $\nu_t\in[0,\nu_{\max}]$;
\item legal-spam proposals fail signal qualification with probability at least $1-\xi$ and are excluded from eligible cost;
\item slashing/penalty mass is nonnegative for $Q_{t,j}=0$ proposals; and
\item honest high-signal proposals arrive with probability at least $p_h>0$ per epoch.
\end{enumerate}
Then $(\nu_t)$ is bounded and admits a nonempty invariant interval $I\subseteq[0,\nu_{\max}]$.
Moreover, there exists $p_\star>0$ such that any honest proposal with margin condition
\[
N_i-\bar\nu C_i \ge \delta>0\quad\text{for all }\bar\nu\in I
\]
is accepted with probability at least $p_\star$.
Hence legal-spam injections cannot drive the system to asymptotic paralysis by exploding $\nu_t$.
\end{theorem}
\begin{proof}[Proof sketch]
Excluded/slashed low-signal proposals do not contribute positive drift to eligible cost.
Projected updates plus bounded clipped costs yield bounded increments, so $(\nu_t)$ remains in a compact interval.
Within that interval, honest proposals with positive utility margin retain strictly positive passing probability under always-valid novelty tests.
\end{proof}

\subsection{Goodhart/Campbell robustness via multi-metric gates}
Novelty metrics can be gamed \citep{Goodhart1975,Campbell1976}.
Multi-metric gates are useful, but their error accounting must be correct.

\begin{proposition}[Correct bounds for ``must pass all'' gates]
Let $A_k$ be the event that a null candidate passes witness test $k$, and suppose $\mathbb{P}(A_k)\le \alpha_k$ for each $k$.
If adoption requires passing \emph{all} $K$ tests, then without dependence assumptions,
$\mathbb{P}(\cap_{k=1}^K A_k)\le \min_k \alpha_k$.
If pass events are independent under the null, then
$\mathbb{P}(\cap_{k=1}^K A_k)\le \prod_{k=1}^K \alpha_k$.
\end{proposition}

\subsection{Two-tier acceptance and paralysis control}
\begin{definition}[Tiered decision rule]
For each candidate, define:
\begin{itemize}[leftmargin=*]
\item \textbf{Critical gates}: safety, provenance integrity, conservativity.
\item \textbf{Value gates}: progress and novelty.
\end{itemize}
If critical gates pass and value evidence is promising but incomplete, the candidate enters $\Prov$ with bounded deployment scope and horizon $H_{\Prov}$.
Full adoption requires value-gate confirmation before horizon expiry; otherwise rollback to the incumbent.
\end{definition}

\begin{proposition}[Paralysis-risk bound under power floor]
Assume each provisional cycle has probability at least $p_\star>0$ of reaching full adoption (implied by effect-size margins and minimum sample allocations).
Then after $M$ independent cycles, the probability of no full adoption is at most $(1-p_\star)^M$.
\end{proposition}


\section{Incorporation Certificates and the Audit-Closed Protocol}

The log is not only a record of outcomes; it is the \emph{sole} adjudicator. To make this operational, we require each adopted update to be accompanied by an \emph{Incorporation Certificate} that enables deterministic replay and third-party verification.

\begin{definition}[Incorporation Certificate]
An Incorporation Certificate $\Cert$ is a structured record appended to the transparency log that contains at least:
\begin{enumerate}[leftmargin=*]
\item \textbf{Genesis anchors:} digests of $\Gen^{(0)}$, $\Gen^{(1)}$, and hardware-trust profile $\Gen^{(\mathrm{HW})}$, including scoring rules, alpha budgets, role-scheduler policy, adaptive-context policy constraints $(\pi_{\min},c_w)$, admissibility rules, and permitted pilot summaries.
\item \textbf{Candidate commitment:} the committed candidate-set identifier $\Cand_t$, the selected candidate ID, and all generator digests and seeds/commitments.
\item \textbf{Interface change:} any expansion $i:\Obs_0\hookrightarrow\Obs_1$, declared post-processing morphisms, and explicit joint-context witnesses used for comparability.
\item \textbf{Coherence graph:} finite acyclic audit-expression graph $G_t$, per-edge budgets $b(\cdot)$, contraction bounds $\bar\eta(\cdot)$, deterministic-depth cap $L_{\det}$, and deterministic budget cap $\epsilon_{\det}$.
\item \textbf{Provenance attestation:} device IDs, firmware/calibration digests, challenge nonces, signatures, and verification outcomes.
\item \textbf{Data digests:} digest of pooled stream $D_{\mathrm{eval},t}$, role-scheduler commitment, and role-partition digests ($D_{\mathrm{fit}},D_{\mathrm{A},t},D_{\mathrm{H},t},D_{\mathrm{P},t},D_{\mathrm{G},t}$ as applicable).
\item \textbf{Gate transcripts:} deterministic transcripts for each gate (safety, integrity, physical coherence, coherence, conservativity, novelty, garbling, progress), including e-process definitions, observed increments, logged propensities for adaptive sampling, stopping times, resulting e-values, anti-griefing novelty-accounting traces, and calibration-update evidence when used.
\item \textbf{Decision:} the audit-closed decision $\Rej/\Quar/\Prov/\Adopt$ and the deterministic rule outputs that produced it.
\item \textbf{Reproducibility contract:} determinism tier, dependency digests, and replay instructions.
\end{enumerate}
\end{definition}

\begin{remark}
The certificate is not a proof of truth; it is proof that acceptance followed Genesis and can be rechecked from the public log alone.
\end{remark}

\begin{algorithm}[H]
\caption{Audit-Closed Epistemic Incorporation (tiered)}
\begin{algorithmic}[1]
\State \textbf{Input:} log prefix $\Log_{0:t-1}$, fixed Genesis modules $\Gen^{(0)},\Gen^{(1)},\Gen^{(\mathrm{HW})}$
\State Commit candidate set $\Cand_t \gets G(\Log_{0:t-1}, D_{\mathrm{fit}})$ before consuming epoch-$t$ evaluation outcomes
\State Build role-assigned streams from pooled $D_{\mathrm{eval},t}$ using committed scheduler $\rho_t$
\For{each candidate $c\in\Cand_t$}
  \State Verify attestation/provenance, static safety, and admissibility checks (fail-closed)
  \If{$c$ proposes an expansion $i$}
    \State Enter quarantine only if quarantine entry conditions pass
    \State Collect pilot data $D_{\mathrm{P},t}$ under strict non-leakage contract
    \State Collect paired holdout $D_{\mathrm{G},t}$ if needed for garbling certificates
  \EndIf
  \State Check contracted coherence budgets on audit graph $G_t$ (Definition~\ref{def:cohbudget})
  \State Run integrity and physical-coherence gates on synchronized role streams
  \State Run conservativity and progress gates on their role streams
  \If{$c$ proposes an expansion $i$}
    \State Run novelty gate on $D_{\mathrm{P},t}$ and optional garbling gate on $D_{\mathrm{G},t}$
  \EndIf
  \State Apply tiered decision rule: $\Rej/\Quar/\Prov/\Adopt$
  \State Emit Incorporation Certificate $\Cert$
  \State Append $\Cert$ and decision to the transparency log
\EndFor
\end{algorithmic}
\end{algorithm}


\section{E-processes, e-values, and Multiple Testing}

\subsection{Sanitized bounded scoring and outlier resilience}
\begin{definition}[Measurement sanitizer]
Let raw device output be $R_i$.
Genesis fixes a deterministic sanitizer
\[
\psi:R\to \mathcal{Y}_{\mathrm{clean}}\cup\{\bot\},
\]
where $\bot$ denotes anomaly/unusable record.
Scientific scores are computed on $\psi(R_i)$ and bounded by construction.
\end{definition}

\begin{definition}[Anomaly indicator channel]
Define $A_i=\mathbbm{1}\{\psi(R_i)=\bot\}$.
Main scientific gates use sanitized outcomes; anomaly rates are tested by a separate integrity gate.
\end{definition}

\begin{remark}
This avoids epoch-wide denial-of-service from isolated outliers.
Outlier pressure is tracked as explicit evidence rather than forcing automatic transcript discard.
\end{remark}

\subsection{Bounded proper scoring rules}
Genesis uses bounded proper rules directly (no post-hoc clipping).

\begin{definition}[Two bounded proper scoring rules]
\begin{itemize}[leftmargin=*]
\item \textbf{Brier score (binary outcome).} For $y\in\{0,1\}$ and predicted probability $p$, $\mathcal{S}_{\mathrm{Brier}}(p;y)=-(p-y)^2\in[-1,0]$.
\item \textbf{Log score on a finite partition.} Fix a measurable partition of outcomes into $B$ bins; predictions are categorical probabilities with a fixed probability floor $p_{\min}>0$ enforced by smoothing (e.g., Dirichlet prior), so $\log p \in[\log p_{\min},0]$.
\end{itemize}
\end{definition}

\begin{remark}
The partition and smoothing rule are part of the observation interface and therefore auditable.
\end{remark}

\begin{definition}[Multi-resolution partition ladder]
For continuous outcomes, Genesis may fix a nested family of finite partitions $(\Pi_r)_{r=0}^R$ (e.g., dyadic bins) together with fixed weights $(w_r)_{r=0}^R$ and a smoothing rule (probability floor) per resolution.
Define the \emph{ladder score} as the weighted sum of discrete log scores on each partition:
\[
\mathcal{S}_{\mathrm{ladder}}(q;y)=\sum_{r=0}^R w_r \log q_{\Pi_r}(\Pi_r(y)),
\]
where $q_{\Pi_r}$ denotes the induced categorical distribution on cells of $\Pi_r$ after smoothing.
\end{definition}

\begin{remark}
The ladder score provides multi-scale sensitivity while preserving bounded increments.
Changing ladder resolution is treated as a Genesis-level update.
\end{remark}

\begin{theorem}[Sample-efficient ladder discrepancy bound]
\label{thm:ladder}
Let $(\Pi_r)_{r=0}^{R}$ be the fixed ladder with weights $w_r\ge0$, $\sum_r w_r=1$, and per-level sample counts $N_r$ (with $\sum_r N_r=N$).
Assume bounded per-level score increments and fixed probability floors.
Then for any $\delta\in(0,1)$, with probability at least $1-\delta$, the ladder discrepancy estimation error is bounded by
\[
\sum_{r=0}^{R} w_r\,c_r\sqrt{\frac{|\Pi_r|+\log((R+1)/\delta)}{N_r}},
\]
for constants $c_r$ determined by boundedness contracts.
Consequently, for balanced allocations $N_r\asymp N$, the rate is
\[
O\!\left(\sqrt{\frac{1}{N}\sum_{r=0}^{R}w_r|\Pi_r|}\right),
\]
and the induced squared-error complexity scales as
\[
O\!\left(\frac{1}{N}\sum_{r=0}^{R}w_r|\Pi_r|\right),
\]
which is typically far smaller than single-resolution fine-mesh complexity $O(|\Pi_R|/N)$.
\end{theorem}
\begin{proof}[Proof sketch]
Apply level-wise concentration for bounded increments and union bound across resolutions.
The weighted sum structure yields additive complexity in effective bin counts rather than worst-case dependence on the finest partition alone.
\end{proof}


\subsection{Explicit mixture e-process for bounded mean tests}
We test composite nulls of the form $H_0:\mathbb{E}[\Delta]\le \tau$ using bounded increments.

\begin{definition}[Normalized increment]
Let $\Delta_i$ be bounded in $[a,b]$ (known by Genesis).
Assume $\tau\in[a,b]$.
Define $X_i=(\Delta_i-\tau)/(b-a)$ so that $X_i\in[-1,1]$.
\end{definition}
\begin{remark}[Boundedness contract]
The bounds $[a,b]$ are enforced at the sanitized-score level.
Raw anomalies are routed to the integrity channel rather than silently clipped or causing implicit success.
\end{remark}

\begin{definition}[Grid-mixture betting e-process]
Fix a grid $\Lambda=\{\lambda_1,\dots,\lambda_m\}\subset(0,1)$ and fixed weights $\pi_j>0$ with $\sum_j\pi_j=1$ (both fixed in Genesis).
Define
\[
E_n^{(\lambda)}=\prod_{i=1}^n (1+\lambda X_i),
\qquad
E_n=\sum_{j=1}^m \pi_j E_n^{(\lambda_j)}.
\]
\end{definition}

\begin{proposition}[Supermartingale property (with tower rule explicit)]
Assume $(X_i)$ is adapted to $(\mathcal{F}_i)$ and satisfies $\mathbb{E}[X_i\mid \mathcal{F}_{i-1}]\le 0$ under $H_0$.
Then for each fixed $\lambda\in(0,1)$,
\[
\mathbb{E}[E_n^{(\lambda)}\mid \mathcal{F}_{n-1}] \le E_{n-1}^{(\lambda)}.
\]
Moreover, because $(\pi_j)$ are fixed in Genesis (non-adaptive), $(E_n)$ is also a supermartingale.
\end{proposition}
\begin{proof}[Proof sketch]
$E_n^{(\lambda)}=E_{n-1}^{(\lambda)}(1+\lambda X_n)$, so
$\mathbb{E}[E_n^{(\lambda)}\mid \mathcal{F}_{n-1}]=E_{n-1}^{(\lambda)}\mathbb{E}[1+\lambda X_n\mid \mathcal{F}_{n-1}]\le E_{n-1}^{(\lambda)}$.
Mixtures of supermartingales with fixed nonnegative weights are supermartingales.
\end{proof}

\begin{theorem}[Optional-stopping-safe acceptance]
\label{thm:optional}
Accept at the first time $\tau$ such that $E_\tau\ge 1/\alpha$.
Then $\mathbb{P}_{H_0}(\sup_n E_n\ge 1/\alpha)\le \alpha$.
\end{theorem}
\begin{proof}
Ville's inequality \citep{Howard2021CS,Shafer2021Betting,Grunwald2024SafeTesting}.
\end{proof}

\subsection{Adaptive contextual sampling with logged propensities}
This section combines logged-propensity correction with always-valid betting martingales in the off-policy sequential setting \citep{Karampatziakis2021OffPolicyCS,WaudbySmithRamdas2024Betting}.
\begin{definition}[Conservative fixed-ratio IPS increment]
Let $X_i$ be the context selected at trial $i$ from logged policy $\pi_i(\cdot\mid\mathcal{F}_{i-1})$ and let $\Delta_i\in[a,b]$ be the bounded progress increment.
Assume $\tau\in[a,b]$.
With ratio bound
\[
\frac{\mathcal{Q}_P(X_i)}{\pi_i(X_i\mid\mathcal{F}_{i-1})}\le c_w,
\]
define
\[
\widetilde X_i
=
\frac{1}{c_w}
\frac{\mathcal{Q}_P(X_i)}{\pi_i(X_i\mid\mathcal{F}_{i-1})}
\frac{\Delta_i-\tau}{b-a}
\in[-1,1].
\]
\end{definition}

\begin{theorem}[Audit-closed optional stopping under adaptive contextual sampling]
\label{thm:adaptive-optional}
Assume logged propensities and the progress null
\[
H_0^{\mathrm{prog}}:\ \mathbb{E}_{X\sim \mathcal{Q}_P}\!\left[\mathbb{E}[\Delta_i-\tau\mid X,\mathcal{F}_{i-1}]\right]\le 0.
\]
Then $\mathbb{E}[\widetilde X_i\mid\mathcal{F}_{i-1}]\le 0$.
Any fixed-grid mixture e-process built from $(\widetilde X_i)$ is a nonnegative supermartingale, so optional-stopping false-acceptance control is identical to Theorem~\ref{thm:optional}.
\end{theorem}
\begin{proof}
By tower rule and logged propensity correction,
\[
\mathbb{E}\!\left[
\frac{\mathcal{Q}_P(X_i)}{\pi_i(X_i\mid\mathcal{F}_{i-1})}
(\Delta_i-\tau)\,\middle|\,\mathcal{F}_{i-1}
\right]
=
\mathbb{E}_{X\sim\mathcal{Q}_P}\!\left[\mathbb{E}[\Delta_i-\tau\mid X,\mathcal{F}_{i-1}]\right]\le 0.
\]
Scaling by positive constant preserves sign.
\end{proof}

\begin{remark}[Worst-case-ratio paralysis]
When $c_w$ is very large, conservative normalization can make increments nearly zero and stall evidence growth.
We therefore add a variance-adaptive rule that retains validity under a stronger context-wise null.
\end{remark}

\begin{definition}[Variance-adaptive predictable betting increment]
Define raw IPS increment
\[
Y_i=
\frac{\mathcal{Q}_P(X_i)}{\pi_i(X_i\mid\mathcal{F}_{i-1})}
\frac{\Delta_i-\tau}{b-a}.
\]
Let $V_{i-1}>0$ be a predictable variance proxy computed from logged history (e.g., exponential moving average of $Y_k^2$ for $k<i$).
For base grid $\Lambda_{\mathrm{va}}=\{\lambda_1,\dots,\lambda_m\}$ and realized weight
\[
w_i=\frac{\mathcal{Q}_P(X_i)}{\pi_i(X_i\mid\mathcal{F}_{i-1})},
\]
set
\[
\lambda_{i,j}=\min\!\left\{\frac{\lambda_j}{\sqrt{V_{i-1}}},\ \frac{1}{2\max(1,w_i)}\right\}.
\]
\end{definition}

\begin{theorem}[Variance-adaptive optional stopping under contextual exploration]
\label{thm:adaptive-var}
Assume the context-wise null
\[
H_{0,\mathrm{ctx}}^{\mathrm{prog}}:\ \mathbb{E}[\Delta_i-\tau\mid X_i,\mathcal{F}_{i-1}]\le 0\quad\text{a.s.}
\]
and let $\mathcal{H}_i=\sigma(\mathcal{F}_{i-1},X_i)$ (context-revealed pre-outcome filtration).
Assume additionally $\tau\in[a,b]$ and a predictable second-moment contract
\[
\mathbb{E}[Y_i^2\mid \mathcal{H}_i]\le V_{i-1}\quad\text{a.s.}
\]
for the same predictable variance proxy $V_{i-1}>0$.
Define
\[
E_n^{(j)}=\prod_{i=1}^{n}\bigl(1+\lambda_{i,j}Y_i\bigr),\qquad
E_n^{\mathrm{va}}=\sum_{j=1}^{m}\pi_j E_n^{(j)},
\]
with fixed nonnegative weights $\pi_j$ summing to one.
Then $(E_n^{\mathrm{va}})$ is a nonnegative supermartingale under $H_{0,\mathrm{ctx}}^{\mathrm{prog}}$, hence optional-stopping acceptance at threshold $1/\alpha$ controls false acceptance at level $\alpha$.
Moreover, for each grid component $j$ and $\mu_i=\mathbb{E}[Y_i\mid\mathcal{H}_i]$,
\[
\mathbb{E}\!\left[\log E_n^{(j)}\right]
\ge
\sum_{i=1}^{n}\mathbb{E}\!\left[\lambda_{i,j}\mu_i-\lambda_{i,j}^2V_{i-1}\right].
\]
Hence growth is governed by realized predictable variance through $\lambda_{i,j}\propto 1/\sqrt{V_{i-1}}$, rather than solely by worst-case global $c_w$.
\end{theorem}
\begin{proof}[Proof sketch]
Given $\lambda_{i,j}\le 1/(2\max(1,w_i))$ and $|Y_i|\le w_i$, each factor is nonnegative.
Under $H_{0,\mathrm{ctx}}^{\mathrm{prog}}$, $\mathbb{E}[Y_i\mid\mathcal{H}_i]\le0$, so
\[
\mathbb{E}[1+\lambda_{i,j}Y_i\mid\mathcal{H}_i]\le 1.
\]
Iterating tower property yields the supermartingale claim.
For growth, use $\log(1+u)\ge u-u^2$ for $u\in[-1/2,\infty)$ with $u=\lambda_{i,j}Y_i$:
\[
\mathbb{E}[\log(1+\lambda_{i,j}Y_i)\mid\mathcal{H}_i]
\ge
\lambda_{i,j}\mu_i-\lambda_{i,j}^2\mathbb{E}[Y_i^2\mid\mathcal{H}_i]
\ge
\lambda_{i,j}\mu_i-\lambda_{i,j}^2V_{i-1}.
\]
Summing over $i$ yields the stated lower bound.
The adaptive betting argument follows the predictable self-normalization pattern in modern betting-based sequential inference \citep{WaudbySmithRamdas2024Betting}.
\end{proof}

\begin{remark}[Doubly robust extension]
Genesis may replace IPS increments by bounded doubly robust pseudo-outcomes.
The same proofs apply if the pseudo-outcome process satisfies the corresponding conditional-mean nonpositivity contract for the chosen null.
\end{remark}

\subsection{Integrity e-process for anomaly-rate control}
\begin{definition}[Integrity null]
With anomaly indicators $(A_i)$ and allowed anomaly rate $q_{\max}$, test
\[
H_0^{\mathrm{int}}:\ \mathbb{E}[A_i\mid\mathcal{F}_{i-1}]\le q_{\max}
\]
using an always-valid e-process.
\end{definition}

\begin{remark}
If $H_0^{\mathrm{int}}$ is rejected, affected devices/channels are quarantined and recertification is required before they can contribute scientific evidence.
\end{remark}

\subsection{Physical-coherence e-process}
\begin{definition}[Physical consistency gate]
For each physical witness $g_m\in\mathcal{G}_{\mathrm{phys}}$, use residual stream $(R_{m,i})$ and an always-valid e-process for $H_{0,m}^{\mathrm{phys}}$.
Apply multiplicity control across $m$.
Failure of any critical physical witness blocks adoption and triggers one of two audited branches:
\begin{enumerate}[leftmargin=*]
\item \textbf{Spoofing branch:} if spoof indicators fire, immediate reject/escalate.
\item \textbf{Aging/sentinel branch:} if spoof indicators do not fire, only a calibration-update candidate (Section 3) is allowed.
Primary-vs-secondary sentinel checks determine whether to recalibrate sensors, maintain sentinels, or freeze calibration pending sentinel replacement.
\end{enumerate}
\end{definition}

\subsection{Batch and stream policies; multiple alpha budgets}
\begin{theorem}[Dependence-free batch FWER control]
For a batch of $m$ candidates with e-values $e_j$ satisfying $\mathbb{E}[e_j]\le 1$ under their nulls, accept $j$ only if $e_j\ge m/\alpha$.
Then $\mathbb{P}(\text{any false acceptance in the batch})\le \alpha$ without dependence assumptions.
\end{theorem}

\begin{theorem}[Stream-level alpha-spending (dependence-free)]
Let $\{\alpha_t\}$ be fixed with $\sum_t \alpha_t\le \alpha$.
At epoch $t$ with $m_t$ candidates, accept only if e-value $\ge m_t/\alpha_t$.
Under the global null, $\mathbb{P}(\exists\ \text{false adoption})\le \alpha$ without dependence assumptions.
\end{theorem}

\begin{remark}[Separate data and alpha budgets]
Genesis allocates explicit gate-wise budgets
\[
\alpha_{\mathrm{tot}}=
\alpha_{\mathrm{cons}}+\alpha_{\mathrm{prog}}+\alpha_{\mathrm{garb}}+\alpha_{\mathrm{nov}}+\alpha_{\mathrm{int}}+\alpha_{\mathrm{safe}}.
\]
Role assignments may be disjoint or cross-fitted, but must be committed and replayable so that no gate consumes outcome-adaptively selected evidence.
\end{remark}

\section{Garbling Certificates as Log-Verifiable Claims}

Blackwell ordering formalizes ``more informative measurements'' \citep{Blackwell1953}.
We use it only when a garbling relation is \emph{certified} from data, with mode selection based on dimensionality and sample support.

\subsection{Low-dimensional TV mode}
\begin{definition}[Kernel distance under a reference $Y_1$ distribution]
Let $K,L:Y_1\rightsquigarrow Y_0$ be kernels.
Fix a reference distribution $\mu$ on $Y_1$ (taken from logged pilot samples).
Define
\[
d_{\TV,\mu}(K,L)=\mathbb{E}_{Y_1\sim \mu}\big[\TV(K(Y_1,\cdot),L(Y_1,\cdot))\big].
\]
\end{definition}

\begin{definition}[Approximate garbling certificate]
Suppose pilot runs produce paired observations $(Y_1,Y_0)$ and a candidate kernel $K:Y_1\rightsquigarrow Y_0$.
A certificate specifies (i) a tolerance $\tau_g$, (ii) a fixed binning of $Y_1$ into finitely many cells, and (iii) a holdout test that upper-bounds
$d_{\TV,\mu}(\mathcal{L}(Y_0\mid Y_1),K)$ by $\tau_g$ using fresh paired holdout data $D_{\mathrm{G},t}$ and e-value control across bins.
\end{definition}

\begin{assumption}[Applicability envelope]
TV-binning mode is used only when minimum per-bin occupancy and bounded effective-dimension conditions are met.
Otherwise this mode is disallowed.
\end{assumption}

\begin{remark}[How the low-dimensional TV bound is audited]
With the fixed binning of $Y_1$, each bin induces an unconditional two-sample comparison:
(i) the empirical distribution of observed $Y_0$ values in the bin, and
(ii) the distribution of $Y_0$ generated by applying the committed kernel $K$ to the observed $Y_1$ values in the same bin using a logged randomness source.
Total variation between finite empirical distributions is computable; an e-process for bounded discrepancies (or a confidence-sequence bound) can upper-bound the true binwise TV with optional-stopping validity.
A global $d_{\TV,\mu}$ bound is then obtained by a weighted sum across bins, with dependence-free error accounting via alpha-spending.
\end{remark}

\subsection{High-dimensional witness-IPM mode}
\begin{definition}[Witness-class discrepancy]
Let $\mathcal{F}_{\mathrm{pub}}$ be a public bounded witness family and let $\mathfrak{F}_{\mathrm{chal}}$ be a large challenge-family generator.
Given challenge seed $\sigma$, define sampled challenge set $\mathcal{F}_{\mathrm{chal}}(\sigma)\subset\mathfrak{F}_{\mathrm{chal}}$.
Set
\[
\mathcal{F}_\sigma=\mathcal{F}_{\mathrm{pub}}\cup \mathcal{F}_{\mathrm{chal}}(\sigma).
\]
All witness functions satisfy $f:Y_0\to[-1,1]$.
Define
\[
d_{\mathcal{F}_\sigma}(P,Q)=\max_{f\in\mathcal{F}_\sigma}\left|\mathbb{E}_{P}[f]-\mathbb{E}_{Q}[f]\right|.
\]
\end{definition}

\begin{definition}[High-dimensional garbling certificate]
A high-dimensional certificate specifies $K$, public witness family commitment, challenge-family generator commitment, tolerance $\tau_{\mathcal{F}}$, and per-witness holdout e-processes testing
\[
H_{0,f}:\ \mathbb{E}\!\left[f(Y_0)-\mathbb{E}_{\tilde Y_0\sim K(Y_1,\cdot)}f(\tilde Y_0)\right]\le \tau_{\mathcal{F}}
\]
for all $f\in\mathcal{F}_\sigma$, with multiplicity correction across witnesses.
Candidate commitment is logged \emph{before} challenge seed reveal to block witness-overfitting attacks.
\end{definition}

\begin{remark}
In high dimension, guarantees are witness-class-relative.
Challenge randomization prevents a fixed public witness set from being the sole optimization target.
Conclusions remain restricted to the audited witness family.
\end{remark}

\begin{proposition}[Challenge-randomization anti-overfitting bound]
Suppose a candidate kernel matches all public witnesses but violates at least a $\beta$ fraction of challenge-family functions by margin larger than $\tau_{\mathcal{F}}$.
If $m$ challenge witnesses are sampled uniformly without replacement after candidate commitment, then the probability that all sampled challenge tests miss the violating region is at most $(1-\beta)^m\le e^{-\beta m}$.
\end{proposition}
\begin{proof}[Proof sketch]
Each sampled challenge witness misses the violating subset with probability at most $(1-\beta)$.
Multiply over draws and apply $1-x\le e^{-x}$.
\end{proof}

\begin{definition}[Stratified challenge generator]
To avoid vanishing global violation mass in high dimension, Genesis may decompose challenge families by resolution/projection tiers:
\[
\mathfrak{F}_{\mathrm{chal}}=\bigcup_{r=0}^{R}\mathfrak{F}^{(r)},
\]
with fixed sampling counts $m_r$ per tier and tier-specific minimum sensitivity contracts $\beta_r$.
\end{definition}

\begin{proposition}[Tiered miss-probability bound]
If tier $r$ witnesses miss a violating candidate with probability at most $(1-\beta_r)$ and draws are independent across sampled witnesses, then
\[
\Pr(\text{all challenge witnesses miss})\le \prod_{r=0}^{R}(1-\beta_r)^{m_r}
\le
\exp\!\left(-\sum_{r=0}^{R}\beta_r m_r\right).
\]
Thus practical detectability is controlled by cumulative tier sensitivity $\sum_r \beta_r m_r$, not by a single global high-dimensional $\beta$.
\end{proposition}

\begin{corollary}[Conditional DPI usage]
If an approximate garbling certificate is accepted with sufficiently small $\tau_g$, then conclusions that rely on the Markov chain $\theta\to Y_1\to Y_0$ may be used as \emph{heuristic support}, but acceptance decisions must still be based on directly tested audit predicates (scores/e-values) rather than on untestable mutual-information statements.
\end{corollary}

\section{Multi-Agent Operation without Infrastructure Privilege}

A multi-agent setting should avoid privileged shared repositories while resisting Sybil manipulation and selective witness disclosure.

\begin{definition}[Permissioned witness registry]
Genesis maintains a permissioned registry $\mathcal{R}_t$ of eligible witness identities with authenticated credentials and per-identity weight caps.
Registry updates require certified governance transitions.
\end{definition}

\begin{definition}[Committee sampling]
For each epoch, a witness committee $\mathcal{W}_t\subseteq\mathcal{R}_t$ is sampled by a public randomness rule (e.g., beacon-seeded VRF).
\end{definition}

\begin{definition}[Liveness quorum and replacement]
Genesis sets quorum thresholds $(\gamma_{\min},q_{\max}^{\mathrm{miss}})$.
If estimated honest-weight lower bound drops below $\gamma_{\min}$ or missing-disclosure rate exceeds $q_{\max}^{\mathrm{miss}}$, the committee is rotated and non-responsive witnesses are penalized/slashed.
\end{definition}

\begin{definition}[Mandatory disclosure and neutral imputation]
Each selected witness must log attested raw data, preprocessing transcript, and replay metadata.
For witness $w$, define
\[
\widetilde E_n^{(w)}=
\begin{cases}
E_n^{(w)}, & \text{if transcript is valid and replayable},\\
1, & \text{otherwise}.
\end{cases}
\]
Aggregate by fixed weights $\omega_w\ge0$, $\sum_w\omega_w=1$:
\[
\bar E_n=\sum_{w\in\mathcal{W}_t}\omega_w\widetilde E_n^{(w)}.
\]
\end{definition}

\begin{theorem}[Sequential validity under disclosure failures]
If each valid witness process $(E_n^{(w)})$ is a nonnegative supermartingale under the corresponding null, then $(\bar E_n)$ is also a nonnegative supermartingale, without independence assumptions.
\end{theorem}
\begin{proof}
Linearity of conditional expectation preserves supermartingale validity.
The imputed value $1$ is a constant martingale and therefore cannot inflate evidence.
\end{proof}

\begin{remark}
Correlation across witnesses does not invalidate the aggregation theorem.
Sybil resistance comes from permissioned membership, identity-weight caps, and public committee sampling.
\end{remark}

\begin{theorem}[Log-growth lower bound under neutral-imputation Byzantine model]
\label{thm:byz-growth}
Assume a fixed gate and an epoch-wise committee with total honest witness weight $\omega_n\ge\gamma>0$ at each $n$.
Let $\widetilde E_n^{\mathrm{hon}}$ denote the honest-weighted mean contribution and assume
\[
\liminf_{n\to\infty}\frac{1}{n}\log \widetilde E_n^{\mathrm{hon}}\ge R>0
\]
almost surely.
Assume all non-honest witnesses contribute neutral-imputed value $1$ (worst-case non-cooperative behavior under disclosure failure).
Then
\[
\liminf_{n\to\infty}\frac{1}{n}\log \bar E_n
\ge R>0.
\]
\end{theorem}
\begin{proof}[Proof sketch]
Write
\[
\bar E_n=(1-\omega_n)\cdot 1+\omega_n\widetilde E_n^{\mathrm{hon}},\qquad \omega_n\ge\gamma.
\]
Since all terms are nonnegative and $\omega_n\ge\gamma$,
\[
\bar E_n\ge \gamma\,\widetilde E_n^{\mathrm{hon}}.
\]
Taking logs and liminf gives
\[
\liminf_{n\to\infty}\frac{1}{n}\log\bar E_n
\ge
\liminf_{n\to\infty}\frac{1}{n}\log\widetilde E_n^{\mathrm{hon}}
+\lim_{n\to\infty}\frac{1}{n}\log\gamma
\;=\;
\liminf_{n\to\infty}\frac{1}{n}\log\widetilde E_n^{\mathrm{hon}}
\;\ge\;R,
\]
since $\lim_{n\to\infty}\frac{1}{n}\log\gamma=0$ for fixed $\gamma>0$.
This proves the claim.
\end{proof}

\begin{corollary}[Asymptotic anti-paralysis implication]
Under Theorem~\ref{thm:byz-growth}, threshold crossing $\bar E_n\ge 1/\alpha$ occurs in finite time almost surely.
An asymptotic epoch scale is
\[
n_\alpha \approx \frac{\log(1/\alpha)+\log(1/\gamma)}{R}.
\]
\end{corollary}

\begin{remark}
If $\gamma=0$ (no honest weight), no growth guarantee is possible.
This is why liveness quorum and committee-replacement constraints are mandatory Genesis components.
\end{remark}

\begin{remark}[Scope of the Byzantine-growth guarantee]
Theorem~\ref{thm:byz-growth} is intentionally scoped to the disclosure-failure threat model where non-honest witnesses are neutral-imputed.
If adversarial witnesses can submit fully valid but systematically anti-evidence streams, additional robust-aggregation contracts are required; the present theorem does not claim coverage for that stronger adversary.
\end{remark}

\section{Genesis Architecture without Hidden Privilege}

Genesis is unavoidably an explicit exogenous commitment.
The claim is not ``no prior commitments''; it is ``no hidden privileged runtime evaluator after commitments are fixed.''

\begin{definition}[Three-module Genesis]
\begin{itemize}[leftmargin=*]
\item $\Gen^{(0)}$ (constitutional): log integrity, admissibility meta-rules, alpha-accounting rules, governance procedure.
\item $\Gen^{(1)}$ (domain): contexts, scores, candidate classes, cost models, power floors.
\item $\Gen^{(\mathrm{HW})}$ (hardware trust profile): attestation formats, key rotation, calibration validity, and recertification policy.
\end{itemize}
\end{definition}

\begin{remark}
Updates to $\Gen^{(1)}$ or $\Gen^{(\mathrm{HW})}$ are treated as certified candidates under $\Gen^{(0)}$.
This keeps governance explicit and auditable without pretending initialization is content-free.
\end{remark}

\section{Observable-Only Semantics and Causality}

Causal claims are treated as hypotheses about \emph{interventional contexts} that can be executed and logged.

\begin{definition}[Interventions as contexts]
A do-operation corresponds to a context in $\Obs$ whose protocol enforces an intervention (e.g., randomized assignment).
Only causal quantities estimable from logged interventional distributions may be used in audit predicates \citep{Pearl2009Causality,Peters2017Elements}.
\end{definition}

\begin{remark}
Graphical structure (DAGs) is an internal representation and is not directly auditable.
Failure modes of invariance-based claims should be tracked explicitly \citep{Arjovsky2019IRM,Rosenfeld2021IRMrisks}.
\end{remark}

\section{Forward Progress and External Validity}

\subsection{Forward progress in observable score}
Error control prevents false acceptances; progress requires a positive signal.
We state a progress guarantee in \emph{observable} terms, avoiding unobservable ``true KL'' references.

\begin{definition}[Progress reference generator]
Genesis fixes a reference generator $\mathcal{Q}_P$ over designated progress contexts.
At epoch $t$, progress-role records are those tagged $(H,\cdot)$ by $\rho_t$.
\end{definition}

\begin{theorem}[Forward progress false-acceptance control (observable form)]
\label{thm:progress}
For candidate $j$ in epoch $t$, let $H^{\mathrm{prog}}_{0,t,j}:\mathbb{E}_{\mathcal{Q}_P}[\Delta_{P,t}^{(j)}]\le 0$.
If progress acceptance requires rejection of $H^{\mathrm{prog}}_{0,t,j}$ at level $\alpha^{\mathrm{prog}}_{t,j}$ and
\[
\sum_t\sum_{j\in\Cand_t}\alpha^{\mathrm{prog}}_{t,j}\le \alpha_{\mathrm{prog}},
\]
then with probability at least $1-\alpha_{\mathrm{prog}}$, no candidate is falsely accepted as progress-positive.
\end{theorem}
\begin{proof}[Proof sketch]
Apply Ville's inequality and union bound over all $(t,j)$.
\end{proof}

\begin{remark}
If $\mathcal{S}$ is log score on a fixed discrete partition, expected score improvement corresponds to improved cross-entropy on that partition.
Any interpretation in terms of latent ``true'' KL is secondary and not used for audit decisions.
\end{remark}

\subsection{Distribution shift with bounded-horizon recovery}
Long-running autonomy faces covariate shift and concept drift \citep{SugiyamaKawanabe2012,Gama2014ConceptDrift}.
Importance weighting and nested expectation estimators can be computationally expensive and numerically fragile; Genesis should therefore treat any such correction as a costed, audited component rather than a free primitive \citep{Rainforth2018NestingMC}.
We formalize an audit-closed response with finite-horizon exit conditions.

\begin{definition}[Audit-closed drift detector]
Genesis fixes a drift statistic $Z_t$ computable from logged data (e.g., score residuals under the currently adopted theory on a fixed calibration context set) and an always-valid e-process $(E^{\mathrm{drift}}_n)$ for the null ``no drift''.
A drift trigger occurs at the first time $\tau$ such that $E^{\mathrm{drift}}_\tau \ge 1/\alpha_{\mathrm{drift}}$, where $\alpha_{\mathrm{drift}}$ is an alpha budget reserved for drift control.
\end{definition}

\begin{definition}[Post-trigger state transition]
Upon a drift trigger, the protocol enters a \emph{drift mode} in which:
(i) non-safety-critical full adoptions are suspended,
(ii) quarantine pilots and provisional safety updates are permitted,
and (iii) within $H_{\max}$ epochs at least one of the following must be proposed:
\begin{enumerate}[leftmargin=*]
\item a re-anchoring or era-transition theory candidate; or
\item a calibration-update candidate for physical witnesses when aging branch conditions hold.
\end{enumerate}
All state transitions are appended to the transparency log and are therefore replayable.
\end{definition}

\begin{definition}[Exogenous-drift certification]
For each triggered epoch, Genesis computes an external-change certificate bit
\[
\chi_t^{\mathrm{ext}}\in\{0,1\},
\]
set to $1$ only if:
\begin{enumerate}[leftmargin=*]
\item the drift trigger is corroborated by physical-coherence evidence from independent witness sets;
\item spoofing indicators remain below their escalation thresholds; and
\item minimum-separation and cooldown rules prevent repeated counting of one unresolved event.
\end{enumerate}
\end{definition}

\begin{definition}[Root continuity debt]
Root Genesis fixes an immutable calibration generator $\mathcal{Q}_{\mathrm{root}}$.
For a candidate $T'$, define
\[
D_{\mathrm{root}}(T')=\max\Bigl\{0,\,-\mathbb{E}_{\mathcal{Q}_{\mathrm{root}}}\big[\mathcal{S}(T';Y)-\mathcal{S}(T;Y)\big]-\delta_{\mathrm{root}}\Bigr\}.
\]
Let
\[
m_t^{\mathrm{ext}}=\sum_{s\le t}\chi_s^{\mathrm{ext}}
\]
be the cumulative number of \emph{externally certified} drift episodes.
For constants $\lambda>0$ and $\kappa_{\min}\in(0,1]$, define
\[
\kappa_t=\max\{\kappa_{\min},e^{-\lambda m_t^{\mathrm{ext}}}\}.
\]
Define effective debt
\[
D_{\mathrm{root}}^{\mathrm{eff}}(T',t)=\kappa_t D_{\mathrm{root}}(T').
\]
Candidates must satisfy $D_{\mathrm{root}}^{\mathrm{eff}}(T',t)\le D_{\max}$.
\end{definition}

\begin{remark}
When physical regimes change irreversibly, $\kappa_t$ decreases and progressively relaxes continuity pressure while preserving an explicit auditable continuity signal.
Because only externally certified episodes affect $m_t^{\mathrm{ext}}$, repeated endogenous drift provocation cannot force arbitrary memory erasure.
\end{remark}

\begin{proposition}[Root-memory erosion resistance]
Under the above definition, any adversarial sequence that triggers drift alarms without external-change certification leaves $m_t^{\mathrm{ext}}$ unchanged and therefore cannot reduce $\kappa_t$.
Moreover, $\kappa_t\ge\kappa_{\min}$ always, so continuity debt never vanishes.
\end{proposition}

\begin{definition}[Calibration-only debt handling]
For calibration-only candidates (no theory-content change), root continuity debt is evaluated on sentinel contexts with the current adopted theory held fixed.
Such candidates are exempt from progress-gate root debt but must satisfy physical witness restoration and conservativity on shared contexts.
\end{definition}

\begin{definition}[Re-anchoring update (mixture form)]
A re-anchoring proposal is a candidate update that replaces the anchor generator $\mathcal{Q}_A$ by
\[
\mathcal{Q}_A'=(1-\rho_t)\mathcal{Q}_A+\rho_t\,\mathcal{Q}_{\mathrm{new}},
\]
where $\rho_t\in(0,1]$ is chosen by a Genesis-fixed severity rule and $\mathcal{Q}_{\mathrm{new}}$ is a proposed generator for restored representativeness.
The proposal is admissible only if it:
\begin{enumerate}[leftmargin=*]
\item is derived from the logged drift evidence (certificate includes the drift e-process transcript);
\item preserves anti-bias, integrity, and data-role invariants from $\Gen^{(0)}$; and
\item satisfies the effective root continuity debt bound.
\end{enumerate}
\end{definition}

\begin{theorem}[Coherence robustness under mixture anchoring]
\label{thm:mix-coh}
For a fixed candidate $T'$ and any audit path $p$ in the certified coherence graph, let $\Delta_p(Y)$ be the path inconsistency random variable under anchor sample $Y$.
Assume
\[
\mathbb{E}_{\mathcal{Q}_A}[\Delta_p]\le \epsilon_{\mathrm{coh}},
\qquad
\mathbb{E}_{\mathcal{Q}_{\mathrm{new}}}[\Delta_p]\le \epsilon_{\mathrm{coh}},
\]
as certified by independent gate transcripts.
Then under mixture anchor
\[
\mathcal{Q}_A'=(1-\rho_t)\mathcal{Q}_A+\rho_t\mathcal{Q}_{\mathrm{new}},
\]
we have
\[
\mathbb{E}_{\mathcal{Q}_A'}[\Delta_p]\le \epsilon_{\mathrm{coh}}
\]
for every certified path $p$.
\end{theorem}
\begin{proof}
Linearity of expectation under mixture distributions gives
\[
\mathbb{E}_{\mathcal{Q}_A'}[\Delta_p]
=(1-\rho_t)\mathbb{E}_{\mathcal{Q}_A}[\Delta_p]+\rho_t\mathbb{E}_{\mathcal{Q}_{\mathrm{new}}}[\Delta_p]
\le \epsilon_{\mathrm{coh}}.
\]
\end{proof}

\begin{definition}[Subgraph-local drift tests]
Let certified audit graph $G_t$ decompose into monitored subgraphs
\[
G_t=\bigcup_{k=1}^{K}G_t^{(k)},
\]
each linked to a disjoint observable interface family.
For each $k$, Genesis defines a local drift e-process
\[
\bigl(E_n^{\mathrm{drift},(k)}\bigr)
\]
for null ``no drift in subgraph $k$'' with local alpha budget $\alpha_k$.
\end{definition}

\begin{theorem}[Audit-closed causal fault localization]
\label{thm:fault-local}
Suppose $\sum_{k=1}^{K}\alpha_k\le \alpha_{\mathrm{drift}}$ and closed-testing acceptance is used for local exemptions.
If global drift triggers but for a subgraph $k$ the local no-drift hypothesis is not rejected, then $G_t^{(k)}$ may be exempted from quarantine/re-anchoring updates while affected subgraphs continue in drift mode.
This localization preserves familywise false-localization control at level $\alpha_{\mathrm{drift}}$.
\end{theorem}
\begin{proof}[Proof sketch]
Each local e-process is always-valid for its null.
Closed testing with alpha allocation $\{\alpha_k\}$ controls familywise error under arbitrary dependence.
We use the classical closed-testing principle in e-value form \citep{Marcus1976ClosedTesting}.
Therefore exempting only nonrejected subgraphs does not inflate global false-localization risk beyond $\alpha_{\mathrm{drift}}$.
\end{proof}

\begin{proposition}[Finite-exit sufficient condition]
Assume there exists an admissible transition candidate with positive progress margin on $\mathcal{Q}_{\mathrm{new}}$, bounded effective root debt, and per-epoch confirmation probability at least $p_{\mathrm{exit}}>0$.
Then the probability of remaining in drift mode for more than $H_{\max}+L$ epochs is at most $(1-p_{\mathrm{exit}})^L$.
\end{proposition}

\section{Synthetic Self-Driving-Lab Demonstration}

We include a fully reproducible synthetic benchmark for availability, false-acceptance behavior, adaptive-sampling validity, and hardware-aging recovery.

\subsection{Setup}
All numbers in this section are generated by a single script (\texttt{appendix\_reproducibility\_code.py}, Appendix C) with fixed seed 20260222.
The manuscript, appendix code, and appendix output manifest are publicly archived at \url{https://doi.org/10.5281/zenodo.18728589}.
The script also performs an automatic manuscript-consistency check over the Section 13 source slice in the canonical manuscript file (\texttt{Audit-Closed AI Scientist Protocol.tex}): every rounded numeric value reported here must appear in that slice, otherwise the check fails.
Gate increments use bounded sanitized scores and mixture e-processes with grid
$\Lambda=\{0.1,0.3,0.5,0.7,0.9\}$ and threshold $1/\alpha=100$ for $\alpha=0.01$.
Per-run sample size is $N=220$ when applicable.

Monte Carlo repetitions:
\begin{itemize}[leftmargin=*]
\item 1500 runs for outlier resilience.
\item 700 runs for drift-mode bounded-horizon exit.
\item 1200 runs for Byzantine witness liveness.
\item 800 runs for novelty legal-spam stress.
\item 2000 runs for adaptive contextual progress testing.
\item 2000 runs for high-$c_w$ IPS-paralysis stress.
\item 1500 runs for root-debt erosion attack stress.
\item 1500 runs for subgraph fault-localization stress.
\item 2500 runs for sentinel-hierarchy stress.
\item 1200 runs for physical recalibration stress.
\end{itemize}

Outlier scenarios:
\begin{itemize}[leftmargin=*]
\item A: $p_{\mathrm{out}}=1.5\%$ per sample.
\item B: $p_{\mathrm{out}}=5\%$ per sample.
\item Null: no true gain, $p_{\mathrm{out}}=1.5\%$.
\end{itemize}

Policies:
\begin{itemize}[leftmargin=*]
\item \textbf{Strict abort}: any out-of-range record aborts the epoch transcript.
\item \textbf{Sanitized}: outliers map to neutral contributions and are tracked by the integrity gate.
\end{itemize}

Additional stress settings:
\begin{itemize}[leftmargin=*]
\item \textbf{Byzantine witness liveness:} honest weight $\gamma=0.35$, neutral-imputed Byzantine weight $0.65$, honest log-growth target $R=0.05$, horizon 400.
\item \textbf{Novelty legal spam:} 12 low-signal high-cost proposals and 2 honest proposals per epoch; baseline dual update vs signal-qualified eligible-cost plus slashing defense.
\item \textbf{Adaptive contextual sampling:} target reference $\mathcal{Q}_P=(0.5,0.3,0.2)$; baseline policy $\pi=(0.5,0.3,0.2)$; adaptive policy $\pi=(0.4,0.25,0.35)$ with bounded ratio $c_w\le1.25$.
\item \textbf{High-$c_w$ IPS stress:} reference $\mathcal{Q}_P=(0.4,0.0667,\dots)$ and exploratory policy $\pi=(0.0004,0.1111,\dots)$ yielding $c_w=1000$; compare fixed-$c_w$ normalization vs variance-adaptive predictable betting.
\item \textbf{Root-debt erosion attack:} repeated adversarial drift alarms with rare true exogenous shifts; compare legacy decay counter vs exogenous-certified counter.
\item \textbf{Fault localization:} two independent subgraphs with drift injected in one branch; compare global-freeze vs localized exemption policy.
\item \textbf{Sentinel hierarchy:} compare single-sentinel calibration against primary/secondary sentinel architecture under simultaneous sensor and sentinel aging.
\item \textbf{Physical recalibration:} benign sensor aging drift with occasional spoof attacks; compare no-recalibration vs audited recalibration branch.
\end{itemize}

\subsection{Results}
\begin{table}[H]
\centering
\small
\begin{tabular}{lrrrr}
\toprule
Scenario & Abort (strict) & Full (strict) & Full (sanitized) & Integrity fail \\
\midrule
A ($p_{\mathrm{out}}=1.5\%$) & 0.966 & 0.021 & 0.642 & 0.000 \\
B ($p_{\mathrm{out}}=5\%$) & 1.000 & 0.000 & 0.598 & 0.000 \\
Null (no gain) & 0.961 & 0.000 & 0.001 & 0.000 \\
\bottomrule
\end{tabular}
\caption{Outlier stress test. Sanitized scoring avoids strict-abort paralysis and remains conservative in the null setting.}
\end{table}

For bounded-horizon drift recovery, exit rate from drift mode is $1.000$, stuck rate is $0.000$, median exit delay is $4$ epochs, and 90th percentile exit delay is $8$ epochs.

\begin{table}[H]
\centering
\small
\begin{tabular}{lrr}
\toprule
Stress test metric & Baseline rule & Defended rule \\
\midrule
Novelty legal-spam: mean final $\nu_t$ & 15.027 & 1.727 \\
Novelty legal-spam: honest acceptance rate & 0.051 & 0.718 \\
\bottomrule
\end{tabular}
\caption{Economic denial-of-service stress test for novelty dual updates. Signal-qualified eligibility plus slashing prevents dual blow-up.}
\end{table}

For Byzantine witness liveness, threshold-crossing rate is $1.000$ (within 400 epochs), median crossing time is $113$ epochs, and 90th percentile crossing time is $149$ epochs.
This matches Theorem~\ref{thm:byz-growth}: with $\gamma=0.35$, $R=0.05$, and $\alpha=0.01$, the scale
\[
\frac{\log(1/\alpha)+\log(1/\gamma)}{R}
=\frac{\log(100)+\log(1/0.35)}{0.05}\approx 113.3
\]
agrees with the observed median crossing epoch.

\begin{table}[H]
\centering
\small
\begin{tabular}{lrr}
\toprule
Adaptive progress metric & Baseline policy & Adaptive policy \\
\midrule
Alternative: threshold-crossing rate & 0.273 & 0.335 \\
Alternative: mean $\log E$ at horizon & 2.682 & 3.406 \\
Null: threshold-crossing rate & 0.0005 & 0.0010 \\
Max ratio $c_w$ & 1.00 & 1.25 \\
\bottomrule
\end{tabular}
\caption{Adaptive contextual sampling with logged propensities and IPS-normalized increments. Evidence growth increases under the adaptive policy while null crossing rates remain near zero.}
\end{table}

\begin{table}[H]
\centering
\small
\begin{tabular}{lrr}
\toprule
High-$c_w$ stress metric & Fixed-$c_w$ IPS & Variance-adaptive IPS \\
\midrule
Alternative: threshold-crossing rate & 0.000 & 0.112 \\
Alternative: mean $\log E$ at horizon & 0.005 & 2.954 \\
Null: threshold-crossing rate & 0.000 & 0.000 \\
\bottomrule
\end{tabular}
\caption{IPS paralysis stress ($c_w=1000$). Conservative fixed normalization stalls evidence growth, while variance-adaptive predictable betting restores progress without null inflation in this benchmark.}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Root-debt attack metric & Legacy decay counter & Exogenous-certified counter \\
\midrule
Median drift count used in decay & 84 & 6 \\
Median $\kappa_t$ & 0.200 & 0.549 \\
Rate of near-forgetfulness ($\kappa_t\le 0.25$) & 1.000 & 0.001 \\
\bottomrule
\end{tabular}
\caption{Root-memory erosion stress. Counting only externally certified drift events prevents adversarial decay acceleration.}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{lr}
\toprule
Fault-localization metric & Value \\
\midrule
Unaffected-subgraph uptime (global freeze) & 0.738 \\
Unaffected-subgraph uptime (localized exemption) & 1.000 \\
False local quarantine rate & 0.0005 \\
\bottomrule
\end{tabular}
\caption{Subgraph-local drift handling preserves unaffected components while maintaining strict localization error control.}
\end{table}

\begin{table}[H]
\centering
\small
\begin{tabular}{lrr}
\toprule
Sentinel-aging metric & Single sentinel & Hierarchical sentinels \\
\midrule
False sensor recalibration rate & 0.433 & 0.001 \\
Operational freeze rate & 0.000 & 0.048 \\
\bottomrule
\end{tabular}
\caption{Primary/secondary sentinel hierarchy prevents calibration collapse loops at small availability cost.}
\end{table}

\begin{table}[H]
\centering
\small
\begin{tabular}{lrr}
\toprule
Physical-coherence metric & No recalib branch & With recalib branch \\
\midrule
Legitimate-aging block rate & 0.394 & 0.038 \\
Recovery rate after aging-induced block & 0.000 & 0.920 \\
Spoof-attack detection rate & 1.000 & 1.000 \\
\bottomrule
\end{tabular}
\caption{Hardware-aging stress test. Audited recalibration sharply improves availability without weakening spoof detection.}
\end{table}

The new stress tests align with the added theory: high-ratio contextual exploration remains auditable with variance-adaptive betting (Theorem~\ref{thm:adaptive-var}), root-memory erosion is bounded by exogenous-only decay updates, and localized drift handling preserves unaffected subgraphs (Theorem~\ref{thm:fault-local}).

\section{Implementation and Reproducibility Commitments}

Each certificate must include reproducibility tier, deterministic dependency digests, seed commitments, and replay instructions.
Tiered replay contracts are essential because physical pipelines combine deterministic software and stochastic hardware.

For this manuscript, complete executable code and exact output manifest are included in Appendix C and Appendix D.
Public archive DOI (full URL): \url{https://doi.org/10.5281/zenodo.18728589}.
Reproduction command:
\begin{verbatim}
python appendix_reproducibility_code.py
\end{verbatim}
The script writes \texttt{appendix\_reproducibility\_results.json}, including:
\begin{itemize}[leftmargin=*]
\item \texttt{reproducibility\_manifest}: UTC generation time, Python/runtime metadata, script SHA-256, and core-results canonical payload SHA-256;
\item \texttt{manuscript\_consistency\_check}: pass/fail status and missing-value list for the manuscript numeric audit.
\end{itemize}
All tables in Section 13 are transcribed directly from that JSON manifest.

\section{Conclusion}

Audit-closed scientific governance must satisfy both formal validity and physical deployability.
The framework presented here combines typed observable semantics, non-vacuous coherence auditing with deterministic-depth controls, optional-stopping-safe inference under both fixed and adaptive contextual sampling, anti-griefing novelty accounting, attested provenance plus physical-coherence checks with hierarchical sentinels, Sybil-resistant federated evidence, and availability-preserving state transitions.
Beyond definitions, it provides system-level guarantees: a variance-adaptive contextual e-process that avoids high-ratio IPS paralysis, a Byzantine-resilient log-growth lower bound for aggregated evidence, self-stabilized novelty dual dynamics under legal-spam pressure, sample-efficient multi-resolution discrepancy auditing, TV--Wasserstein translation for deterministic scientific abstractions, root-debt erosion resistance via exogenous certification, and fault-localized drift recovery with familywise error control.
Under explicit assumptions, it provides replayable update decisions that remain robust to outliers, adversarial infrastructure behavior, and distribution shift.


\appendix

\section{Minimal Incorporation Certificate Schema (informative)}
The following JSON-like skeleton illustrates the minimum fields expected in an Incorporation Certificate.
Genesis fixes the exact canonicalization and hashing rules.

\begin{verbatim}
{
  "genesis": {"G0": "<hash>", "G1": "<hash>", "GHW": "<hash>"},
  "epoch": t,
  "candidate_set": {"id": "<Cand_t>", "commitment": "<hash>"},
  "selected_candidate": "<cid>",
  "interface": {
    "expansion": {"i": "<spec-hash>", "audit_graph": "<hash>"},
    "witnesses": [
      {"o": "...", "m": "...", "o_prime": "...", "a": "<hash>", "b": "<hash>"}
    ],
    "garbling_challenge_seed": "<beacon-seed>"
  },
  "provenance": {
    "attestation": {
      "device_id": "...",
      "firmware": "<hash>",
      "calibration": "<hash>",
      "nonce": "<beacon-derived>",
      "signature": "<sig>"
    }
  },
  "data": {
    "D_eval": "<digest>",
    "role_scheduler": "<hash>",
    "propensity_log_digest": "<digest>",
    "D_fit": "<digest>",
    "D_A": {"digest": "<digest>", "alpha": "alpha_cons"},
    "D_H": {"digest": "<digest>", "alpha": "alpha_prog"},
    "D_P": {"digest": "<digest>", "alpha": "alpha_nov"},
    "D_G": {"digest": "<digest>", "alpha": "alpha_garb"}
  },
  "gates": {
    "coherence": {
      "eps_coh": "eps_coh",
      "eps_det": "eps_det",
      "L_det": "L_det",
      "path_bounds": [...]
    },
    "safety": {...},
    "integrity": {"e_process": {...}, "e_value": ...},
    "physical_coherence": {
      "witnesses": [...],
      "e_values": [...],
      "branch": "spoofing|aging|sentinel_maintenance",
      "recalibration": {
        "sentinel_primary_digest": "<digest>",
        "sentinel_secondary_digest": "<digest>",
        "theta_old": "<hash>",
        "theta_new": "<hash>",
        "threshold_refresh": [...]
      }
    },
    "conservativity": {"e_process": {...}, "e_value": ...},
    "progress": {
      "reference_QP": "...",
      "ips_ratio_cap": "c_w",
      "mode": "fixed_ips|variance_adaptive",
      "variance_trace_digest": "<digest>",
      "e_process": {...},
      "e_value": ...
    },
    "drift": {
      "global_e_process": {...},
      "local_subgraph_e_values": [...],
      "external_cert_bit": "0|1",
      "m_ext": "...",
      "kappa_t": "..."
    },
    "novelty": {
      "dual_nu": "...",
      "eligible_cost_trace": [...],
      "slashing_events": [...],
      "e_process": {...},
      "e_value": ...
    },
    "garbling": {
      "mode": "tv|witness",
      "challenge_set": "...",
      "e_process": {...}
    }
  },
  "decision": "ADOPT|PROVISIONAL|QUARANTINE|REJECT",
  "replay": {"tier": "...", "deps": "<digest>", "seed": "..."}
}
\end{verbatim}

\section{Determinism tiers (informative)}
Genesis should define explicit replay tiers, from ``bitwise deterministic'' to ``statistical replay'' with confidence bounds.
The tier is part of the certificate because reproducibility cost is a first-class constraint.

\section{Complete Reproducibility Code (normative)}
The following file is included verbatim so readers can reproduce every empirical number in Section 13 exactly:
\begin{verbatim}
appendix_reproducibility_code.py
\end{verbatim}
\verbatiminput{appendix_reproducibility_code.py}

\section{Reproducibility Output Manifest (normative)}
Running Appendix C code emits the following machine-readable output:
\begin{verbatim}
appendix_reproducibility_results.json
\end{verbatim}
\verbatiminput{appendix_reproducibility_results.json}


\begin{thebibliography}{99}

\bibitem[Arjovsky et~al.(2019)]{Arjovsky2019IRM}
M.~Arjovsky, L.~Bottou, I.~Gulrajani, and D.~Lopez-Paz.
\newblock Invariant risk minimization.
\newblock arXiv:1907.02893, 2019.
\newblock \url{https://arxiv.org/abs/1907.02893}.

\bibitem[Blackwell(1953)]{Blackwell1953}
D.~Blackwell.
\newblock Equivalent comparisons of experiments.
\newblock {\em The Annals of Mathematical Statistics}, 24(2):265--272, 1953.

\bibitem[Brunton et~al.(2016)]{Brunton2016SINDy}
S.~L. Brunton, J.~L. Proctor, and J.~N. Kutz.
\newblock Discovering governing equations from data by sparse identification of nonlinear dynamical systems.
\newblock {\em PNAS}, 113(15):3932--3937, 2016. doi:10.1073/pnas.1517384113.

\bibitem[Campbell(1976)]{Campbell1976}
D.~T. Campbell.
\newblock Assessing the impact of planned social change.
\newblock Dartmouth College, Public Affairs Center (Occasional Paper Series), 1976.

\bibitem[Chaloner and Verdinelli(1995)]{ChalonerVerdinelli1995}
K.~Chaloner and I.~Verdinelli.
\newblock Bayesian experimental design: a review.
\newblock {\em Statistical Science}, 10(3):273--304, 1995.

\bibitem[Chuat et~al.(2015)]{Chuat2015Gossip}
L.~Chuat, P.~Szalachowski, A.~Perrig, B.~Laurie, and E.~Messeri.
\newblock Efficient gossip protocols for verifying the consistency of certificate logs.
\newblock In {\em Proceedings of the IEEE Conference on Communications and Network Security (CNS)}, 2015.
\newblock doi:10.1109/CNS.2015.7346853.
\newblock Unpaywalled version: \url{https://netsec.ethz.ch/publications/papers/gossip2015.pdf}. 

\bibitem[Cuomo et~al.(2022)]{Cuomo2022PINNReview}
S.~Cuomo et~al.
\newblock Scientific machine learning through physics--informed neural networks: where we are and what's next.
\newblock {\em Journal of Scientific Computing}, 92:88, 2022.

\bibitem[Foster et~al.(2019)]{Foster2019vOED}
A.~Foster, D.~Jankowiak, E.~Bingham, P.~Horsfall, Y.~W. Teh, T.~Rainforth, and N.~Goodman.
\newblock Variational Bayesian optimal experimental design.
\newblock In {\em NeurIPS}, 2019.
\newblock \url{https://papers.nips.cc/paper_files/paper/2019/file/d55cbf210f175f4a37916eafe6c04f0d-Paper.pdf}.

\bibitem[Gama et~al.(2014)]{Gama2014ConceptDrift}
J.~Gama, I.~Zliobait{\.e}, A.~Bifet, M.~Pechenizkiy, and A.~Bouchachia.
\newblock A survey on concept drift adaptation.
\newblock {\em ACM Computing Surveys}, 46(4):44, 2014. doi:10.1145/2523813.

\bibitem[Goguen and Burstall(1992)]{GoguenBurstall1992}
J.~A. Goguen and R.~M. Burstall.
\newblock Institutions: abstract model theory for specification and programming.
\newblock {\em JACM}, 39(1):95--146, 1992.

\bibitem[Goodhart(1975)]{Goodhart1975}
C.~A.~E. Goodhart.
\newblock Problems of monetary management: the {UK} experience.
\newblock Reserve Bank of Australia, Papers in Monetary Economics, Vol.~I, 1975.

\bibitem[Gr{\"u}nwald et~al.(2024)]{Grunwald2024SafeTesting}
P.~Gr{\"u}nwald, R.~de~Heide, and W.~Koolen.
\newblock Safe testing.
\newblock {\em JRSS-B}, 86(5):1091--1128, 2024. doi:10.1093/jrsssb/qkae011.

\bibitem[Howard et~al.(2021)]{Howard2021CS}
S.~R. Howard, A.~Ramdas, J.~McAuliffe, and J.~Sekhon.
\newblock Time-uniform, nonparametric, nonasymptotic confidence sequences.
\newblock {\em Annals of Statistics}, 49(2):1055--1080, 2021. doi:10.1214/20-AOS1991.

\bibitem[Ivanova et~al.(2021)]{Ivanova2021iDAD}
D.~R. Ivanova, A.~Foster, S.~Kleinegesse, M.~U. Gutmann, and T.~Rainforth.
\newblock Implicit deep adaptive design: policy-based experimental design without likelihoods.
\newblock In {\em NeurIPS}, 2021.
\newblock \url{https://papers.neurips.cc/paper_files/paper/2021/hash/d811406316b669ad3d370d78b51b1d2e-Abstract.html}.

\bibitem[Johari et~al.(2017)]{Johari2017Peeking}
R.~Johari, P.~Koomen, L.~Pekelis, and D.~Walsh.
\newblock Peeking at {A/B} tests: why it matters, and what to do about it.
\newblock In {\em KDD}, 2017. doi:10.1145/3097983.3097992.

\bibitem[Johari et~al.(2022)]{Johari2022AlwaysValid}
R.~Johari, L.~Pekelis, and D.~Walsh.
\newblock Always valid inference: continuous monitoring of {A/B} tests.
\newblock {\em Operations Research}, 70(3):1807--1828, 2022. doi:10.1287/opre.2021.2135.

\bibitem[Karniadakis et~al.(2021)]{Karniadakis2021PIML}
G.~E. Karniadakis et~al.
\newblock Physics-informed machine learning.
\newblock {\em Nature Reviews Physics}, 3(6):422--440, 2021. doi:10.1038/s42254-021-00314-5.

\bibitem[Karampatziakis et~al.(2021)]{Karampatziakis2021OffPolicyCS}
Y.~Karampatziakis, A.~Basu, M.~Mineiro, and A.~Dubey.
\newblock Off-policy confidence sequences.
\newblock arXiv:2102.09095, 2021.
\newblock \url{https://arxiv.org/abs/2102.09095}.

\bibitem[King et~al.(2009)]{King2009Automation}
R.~D. King et~al.
\newblock The automation of science.
\newblock {\em Science}, 324(5923):85--89, 2009. doi:10.1126/science.1165620.

\bibitem[Kelsey et~al.(2019)]{NISTIR8213draft}
J.~Kelsey, L.~T.~A.~N. Brand{\~a}o, R.~Peralta, and H.~Booth.
\newblock A reference for randomness beacons: format and protocol version 2 (Draft {NISTIR} 8213).
\newblock NIST, 2019. doi:10.6028/NIST.IR.8213-draft.
\newblock \url{https://nvlpubs.nist.gov/nistpubs/ir/2019/NIST.IR.8213-draft.pdf}.

\bibitem[NIST(2026)]{NISTBeaconProject}
National Institute of Standards and Technology.
\newblock Interoperable randomness beacons project page (status notes indicate the reference document is still a draft).
\newblock Accessed 2026-02-22.
\newblock \url{https://csrc.nist.gov/projects/interoperable-randomness-beacons}.

\bibitem[Lindley(1956)]{Lindley1956}
D.~V. Lindley.
\newblock On a measure of the information provided by an experiment.
\newblock {\em Annals of Mathematical Statistics}, 27(4):986--1005, 1956. doi:10.1214/aoms/1177728069.

\bibitem[Lu et~al.(2024)]{Lu2024AIScientist}
C.~Lu et~al.
\newblock The AI Scientist: Towards fully automated open-ended scientific discovery.
\newblock arXiv:2408.06292, 2024.
\newblock \url{https://arxiv.org/abs/2408.06292}.

\bibitem[Marcus et~al.(1976)]{Marcus1976ClosedTesting}
R.~Marcus, E.~Peritz, and K.~R. Gabriel.
\newblock On closed testing procedures with special reference to ordered analysis of variance.
\newblock {\em Biometrika}, 63(3):655--660, 1976. doi:10.1093/biomet/63.3.655.

\bibitem[Oxford et~al.(2020)]{Oxford2020CTGossip}
M.~Oxford, A.~Malacaria, and C.~Hankin.
\newblock Quantitative verification of certificate transparency gossip protocols.
\newblock In {\em SecureComm}, 2020.
\newblock \url{https://www.prismmodelchecker.org/papers/spc20.pdf}.

\bibitem[Pearl(2009)]{Pearl2009Causality}
J.~Pearl.
\newblock {\em Causality: Models, Reasoning, and Inference}.
\newblock Cambridge University Press, 2nd edition, 2009.

\bibitem[Peng(2011)]{Peng2011}
R.~D. Peng.
\newblock Reproducible research in computational science.
\newblock {\em Science}, 334(6060):1226--1227, 2011. doi:10.1126/science.1213847.

\bibitem[Peters et~al.(2017)]{Peters2017Elements}
J.~Peters, D.~Janzing, and B.~Sch{\"o}lkopf.
\newblock {\em Elements of Causal Inference: Foundations and Learning Algorithms}.
\newblock MIT Press, 2017.

\bibitem[Popper(1959)]{Popper1959}
K.~Popper.
\newblock {\em The Logic of Scientific Discovery}.
\newblock Basic Books, 1959.


\bibitem[Rainforth et~al.(2018)]{Rainforth2018NestingMC}
T.~Rainforth, R.~Cornish, H.~Yang, A.~Warrington, and F.~Wood.
\newblock On nesting Monte Carlo estimators.
\newblock In {\em ICML}, 2018.
\newblock \url{https://proceedings.mlr.press/v80/rainforth18a/rainforth18a.pdf}.

\bibitem[Quine(1951)]{Quine1951}
W.~V.~O. Quine.
\newblock Two dogmas of empiricism.
\newblock {\em The Philosophical Review}, 60(1):20--43, 1951. doi:10.2307/2181906.

\bibitem[Raissi et~al.(2019)]{Raissi2019PINN}
M.~Raissi, P.~Perdikaris, and G.~E. Karniadakis.
\newblock Physics-informed neural networks.
\newblock {\em Journal of Computational Physics}, 378:686--707, 2019. doi:10.1016/j.jcp.2018.10.045.

\bibitem[RFC6962(2013)]{RFC6962}
B.~Laurie, A.~Langley, and E.~Kasper.
\newblock RFC 6962: Certificate transparency.
\newblock RFC Editor, 2013.
\newblock \url{https://www.rfc-editor.org/rfc/rfc6962.html}.

\bibitem[Rosenfeld et~al.(2021)]{Rosenfeld2021IRMrisks}
E.~Rosenfeld, P.~K. Ravikumar, and A.~Rakhlin.
\newblock The risks of invariant risk minimization.
\newblock In {\em ICLR}, 2021.
\newblock \url{https://openreview.net/pdf?id=BbNIbVPJ-42}.

\bibitem[Sandve et~al.(2013)]{Sandve2013}
G.~K. Sandve et~al.
\newblock Ten simple rules for reproducible computational research.
\newblock {\em PLOS Computational Biology}, 9(10):e1003285, 2013. doi:10.1371/journal.pcbi.1003285.

\bibitem[Schmidt and Lipson(2009)]{SchmidtLipson2009}
M.~Schmidt and H.~Lipson.
\newblock Distilling free-form natural laws from experimental data.
\newblock {\em Science}, 324(5923):81--85, 2009. doi:10.1126/science.1165893.

\bibitem[Shafer(2021)]{Shafer2021Betting}
G.~Shafer.
\newblock Testing by betting: a strategy for statistical and scientific communication.
\newblock {\em Journal of the Royal Statistical Society Series A}, 184(2):407--431, 2021. doi:10.1111/rssa.12647.

\bibitem[Sparkes et~al.(2010)]{Sparkes2010RobotScientists}
A.~C. Sparkes et~al.
\newblock Towards robot scientists for autonomous scientific discovery.
\newblock {\em Automated Experimentation}, 2(1):1, 2010. doi:10.1186/1759-4499-2-1.

\bibitem[Sugiyama and Kawanabe(2012)]{SugiyamaKawanabe2012}
M.~Sugiyama and M.~Kawanabe.
\newblock {\em Machine Learning in Non-Stationary Environments: Introduction to Covariate Shift Adaptation}.
\newblock MIT Press, 2012.

\bibitem[Tom et~al.(2024)]{Tom2024SDLchemrev}
G.~Tom et~al.
\newblock Self-driving laboratories for chemistry and materials science.
\newblock {\em Chemical Reviews}, 124(16):9633--9732, 2024. doi:10.1021/acs.chemrev.4c00055.

\bibitem[Tobias and Wahab(2025)]{TobiasWahab2025SDLReview}
A.~V. Tobias and A.~Wahab.
\newblock Autonomous self-driving laboratories: a review.
\newblock {\em Royal Society Open Science}, 12(7):250646, 2025. doi:10.1098/rsos.250646.

\bibitem[Udrescu and Tegmark(2020)]{UdrescuTegmark2020}
S.-M. Udrescu and M.~Tegmark.
\newblock AI Feynman.
\newblock {\em Science Advances}, 6(16):eaay2631, 2020. doi:10.1126/sciadv.aay2631.

\bibitem[Villani(2009)]{Villani2009OT}
C.~Villani.
\newblock {\em Optimal Transport: Old and New}.
\newblock Springer, 2009. doi:10.1007/978-3-540-71050-9.

\bibitem[Waudby-Smith and Ramdas(2024)]{WaudbySmithRamdas2024Betting}
I.~Waudby-Smith and A.~Ramdas.
\newblock Estimating means of bounded random variables by betting.
\newblock {\em Journal of the Royal Statistical Society Series B}, 86(5), 2024.
\newblock doi:10.1093/jrsssb/qkad063.

\bibitem[Yamada et~al.(2025)]{Yamada2025AIScientistV2}
H.~Yamada et~al.
\newblock The AI Scientist-v2: Workshop-level automated scientific discovery via agentic tree search.
\newblock arXiv:2504.08066, 2025.
\newblock \url{https://arxiv.org/abs/2504.08066}.

\end{thebibliography}

\end{document}